---
# INSTRUCTOR SOLUTION — NOT FOR DISTRIBUTION
title: "Trade and Development: An Exploratory Report"
author: "Jane Smith"
date: today
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
  docx:
    toc: true
    number-sections: true
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
theme_set(theme_minimal(base_size = 13))
```

## Introduction

International trade is often described as an engine of economic development.
Yet the relationship between trade openness and outcomes such as income growth
or life expectancy is far from straightforward. As @BekezKezdi2021 argue,
careful empirical analysis requires controlling for confounding variables before
drawing causal conclusions.

This report provides a brief exploratory overview of the relationship between
economic development and health outcomes across countries, drawing on the
framework outlined in @Preston1975.

## Data

The analysis uses a small illustrative dataset of nine countries spanning a wide
range of income levels. The key variables are GDP per capita (USD, PPP-adjusted)
and life expectancy at birth (years). In a full analysis, this would be replaced
by the World Bank World Development Indicators dataset (182 countries, 2017).

```{r}
#| label: data

example_data <- tibble(
  country         = c("Nigeria", "India", "Brazil", "China",
                      "Mexico", "Poland", "Germany", "USA", "Norway"),
  gdp_per_capita  = c(5300, 6600, 14100, 16800,
                      19900, 31900, 53700, 63500, 67300),
  life_expectancy = c(54, 69, 75, 76, 75, 78, 81, 78, 82)
)
```

## Results

### Descriptive overview

```{r}
#| label: fig-scatter
#| fig-cap: "Life expectancy and GDP per capita across nine countries. The log scale on the x-axis reveals a roughly linear relationship."

ggplot(example_data, aes(x = gdp_per_capita, y = life_expectancy)) +
  geom_point(colour = "#00395B", size = 3.5, alpha = 0.85) +
  geom_text(
    aes(label = country),
    nudge_y = 0.7, size = 3.2, colour = "#6F6F6F"
  ) +
  scale_x_log10(
    labels = scales::label_dollar(scale = 1/1000, suffix = "k")
  ) +
  labs(
    x        = "GDP per capita (log scale, USD PPP)",
    y        = "Life expectancy (years)",
    title    = "Higher income is associated with longer life",
    subtitle = "Relationship is non-linear on a linear scale"
  )
```

As shown in @fig-scatter, there is a clear positive association between income
and life expectancy. The relationship flattens at higher income levels — a
pattern first documented by @Preston1975.

## Conclusion

Even a simple cross-country comparison reveals that income and health outcomes
are strongly associated. This raises the question of causation: does higher
income cause better health, or do healthier populations generate more income?
Answering this requires the causal inference tools introduced in Session 5.

## References
