---
title: "Modelling Binary and Categorical Outcomes"
subtitle: "Session 3 · Quantitative Data Analysis"
author: "Claudius Gräbner-Radkowitsch"
date: "2026-04-30"
format:
  revealjs:
    theme: [default, ../../assets/custom.scss]
    slide-number: true
    progress: true
    code-line-numbers: false
    highlight-style: github
    fig-align: center
    self-contained: true
    width: 1280
    height: 720
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(modelsummary)
library(marginaleffects)
library(broom)
library(ggpubr)
```

## Today's session {.center}

::: {.incremental}
1. **Debrief** — Take-home Task 2 *(30 min)*
2. **Why OLS fails** for binary outcomes *(20 min)*
3. **Logistic regression** — intuition and interpretation *(20 min)*
4. **Live demo** with firm exit data *(45 min)*
5. **Your turn** — exercise *(40 min)*
6. **Debrief + Quarto skill** *(15 min)*
:::


# Debriefing the take home

## Take-home Task 2: What we saw

**Common strengths ✓**

- TBD

**Common issues to fix**

- TBD

**One question to reflect on:**  

*You reported that education increases wage by 1,200 EUR. Is that a big effect?*

# Binary outcome variables

## The setup: binary outcomes in business 

Many important business and economics questions have a **yes/no answer:**

| Question | Outcome |
|---|---|
| Will this firm exit the market? | exit = 0 / 1 |
| Did this customer churn? | churn = 0 / 1 |
| Did a loan default? | default = 0 / 1 |
| Was the job application successful? | hired = 0 / 1 |

. . .

::: {.fragment}
What we really want to model is the **probability** of the event:

$$P(\text{exit} = 1 \mid \text{firm size, profit, age, } \ldots)$$
:::

. . .

::: {.fragment}
The natural question: **can we just use OLS?**
:::

---

## What OLS gives us — and why that's a problem

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
set.seed(42)
n <- 300
x <- rnorm(n, mean = 5, sd = 2)
p_true <- plogis(-2 + 0.7 * x)
y <- rbinom(n, 1, p_true)
df <- data.frame(x = x, y = y)

p1 <- ggplot(df, aes(x, y)) +
  geom_jitter(height = 0.03, alpha = 0.4, color = "#2c7bb6") +
  geom_smooth(method = "lm", color = "#d7191c", se = FALSE, linewidth = 1.2) +
  geom_hline(yintercept = c(0, 1), linetype = "dashed", color = "grey40") +
  annotate("label", x = 1.5, y = -0.15, label = "Impossible!\nP < 0", 
           color = "#d7191c", size = 3.5) +
  annotate("label", x = 8.5, y = 1.15, label = "Impossible!\nP > 1", 
           color = "#d7191c", size = 3.5) +
  labs(title = "OLS (Linear Probability Model)",
       x = "Predictor (e.g. log sales)", y = "P(exit = 1)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(df, aes(x, y)) +
  geom_jitter(height = 0.03, alpha = 0.4, color = "#2c7bb6") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "#1a9641", se = FALSE, linewidth = 1.2) +
  geom_hline(yintercept = c(0, 1), linetype = "dashed", color = "grey40") +
  labs(title = "Logistic Regression",
       x = "Predictor (e.g. log sales)", y = "P(exit = 1)") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

ggarrange(p1, p2, ncol = 2)
```

. . . 

::: {.callout-warning}
OLS predictions can fall **outside [0, 1]** — mathematically impossible for probabilities.
:::

---

## Two concrete problems with OLS on binary Y 

::: columns
::: {.column width="50%"}
**Problem 1: Out-of-bounds predictions**

- OLS fits a line → for extreme X values, $\hat{Y} < 0$ or $\hat{Y} > 1$
- A "probability" of −0.12 or 1.34 is meaningless

:::
::: {.column width="50%"}
**Problem 2: Heteroskedastic errors by construction**

- If Y ∈ {0,1}, the variance of ε depends on X:
$$\text{Var}(\varepsilon | x) = P(x)(1 - P(x))$$
- OLS standard errors are invalid → inference is unreliable

:::
:::

. . . 

::: {.fragment}
::: {.callout-note}
**Historical note:** Economists often *do* use OLS on binary outcomes — the "Linear Probability Model" (LPM). It's a useful benchmark, especially with panel data. But you should know its limits.
:::
:::

---

## The solution: the logistic (sigmoid) function

We need a function that maps **any real number → a probability in [0, 1]**

$$\sigma(z) = \frac{e^z}{1 + e^z} = \frac{1}{1 + e^{-z}}$$

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.5
z <- seq(-6, 6, length.out = 200)
ggplot(data.frame(z = z, sigma = plogis(z)), aes(z, sigma)) +
  geom_line(color = "#1a9641", linewidth = 1.5) +
  geom_hline(yintercept = c(0, 0.5, 1), linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  annotate("label", x = 5, y = 0.5, label = "σ(0) = 0.5", size = 4) +
  labs(x = "z (log-odds)", y = "σ(z) = probability",
       title = "The logistic function: maps any input to [0, 1]") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

::: {.fragment}
→ We use this as our link: model $z = \beta_0 + \beta_1 x_1 + \ldots$ and then $\hat{P} = \sigma(z)$
:::

---

## The logit model: what we actually estimate 

Starting from $P = \sigma(\mathbf{X}\boldsymbol{\beta})$, rearranging gives:

$$\ln\left(\frac{P}{1-P}\right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots$$

The left-hand side is the **log-odds** (also called the *logit*):

| Quantity | Formula | Intuition |
|---|---|---|
| Probability | $P$ | Chance of exit (0–1) |
| Odds | $\frac{P}{1-P}$ | "3:1 in favour" |
| Log-odds | $\ln\left(\frac{P}{1-P}\right)$ | What regression models |

::: {.fragment}
**Example:** P(exit) = 0.75 → Odds = 3 → Log-odds = ln(3) ≈ 1.10
:::

::: {.fragment}
**Estimation:** Not OLS, but **maximum likelihood** — find the β that makes the observed data most probable. R does this automatically with `glm()`.
:::

---

## Interpreting logit coefficients: three levels 

Suppose we get $\hat\beta_{\text{profit\_margin}} = -0.43$ from our firm exit model.

. . .

**Level 1 — Log-odds (raw output)**  
A 1-unit increase in profit margin → log-odds of exit decrease by 0.43  
✓ Direction is clear. ✗ Magnitude means nothing to most people.

. . . 

**Level 2 — Odds ratio** $= e^{\hat\beta} = e^{-0.43} \approx 0.65$  
A 1-unit increase in profit margin → odds of exit are multiplied by 0.65 (a 35% reduction)  
✓ Usable. ✗ "Odds" still confuse non-statisticians.

. . .

**Level 3 — Predicted probabilities (most useful for business)**  
"A firm with median characteristics and a 10% profit margin has a 22% chance of exit.  
If that margin drops to −5%, the probability rises to 38%."  
✓ Immediately actionable. ✓ Works for non-technical audiences.

. . .

::: {.fragment}
::: {.callout-tip}
**Rule of thumb:** Always translate your findings into predicted probabilities for a business audience.
:::
:::

---

## How to compute predicted probabilities in R

```{r}
#| eval: false
# Fit the model
model <- glm(exit ~ profit_margin + sales_log + employees_log,
             family = binomial,
             data = firms)

# Predicted probabilities for observed data
firms$pred_prob <- predict(model, type = "response")

# Predicted probability for a specific scenario
new_firm <- data.frame(
  profit_margin = c(0.10, -0.05),   # profitable vs loss-making
  sales_log     = c(10, 10),        # same size
  employees_log = c(3, 3)           # same size
)

predict(model, newdata = new_firm, type = "response")
```

. . . 

::: {.callout-note}
`type = "response"` gives probabilities. `type = "link"` gives log-odds (the default — easy mistake!).
:::

---

## Quarto skill this session: inline R code

Instead of writing "the odds ratio is 0.65" manually, embed the computation:

```markdown
The odds ratio for profit margin is
`{{r}} round(exp(coef(model)["profit_margin"]), 2)`.
```

::: {.fragment}
This renders as: *"The odds ratio for profit margin is 0.65."*

Why does this matter?

- If you update your model or data → **the number updates automatically**
- No copy-paste errors between your model and your text
- Reproducible, professional output
:::

---

## Live demo {.center}

*Switching to `demo.qmd`*

**We will:**

1. Explore the `bisnode-firms` dataset
2. Fit a Linear Probability Model as a benchmark
3. Fit a logistic regression
4. Interpret: log-odds → odds ratios → predicted probabilities
5. Compare LPM and logit side-by-side with `modelsummary()`
6. Write inline R code to report a key finding

---

## Your turn: exercise {.center}

*Open `exercise.qmd` from GitHub Classroom*

**You will:**

- Extend the logistic model with additional predictors
- Interpret your coefficients (all three levels)
- Compute predicted probabilities for two contrasting firm profiles
- Write one paragraph reporting your key finding using **inline R code**

*Work individually or in pairs. I'll circulate. ~40 minutes.*

---

## Key takeaways 

::: {.incremental}
- When Y is binary, we model **P(Y=1|X)** — not Y directly
- **OLS fails** because it produces out-of-bounds predictions and invalid standard errors
- **Logistic regression** uses the sigmoid function to guarantee $\hat{P} \in [0,1]$
- Coefficients are in **log-odds space** — always translate to probabilities for interpretation
- Use `glm(..., family = binomial)` in R; use `predict(..., type = "response")` for probabilities
- **Inline R code** makes your results reproducible and error-proof
:::

::: {.fragment}
**Next session (May 7):** *What can go wrong — biases and diagnostics*  
We'll look at multicollinearity, heteroskedasticity, influential observations, and how to detect them.
:::
