---
title: "Using AI Tools"
subtitle: "Advanced Visualizations as an Example"
author: "Prof. Dr. Claudius Gräbner-Radkowitsch"
institute: "Europa-Universität Flensburg"
date: "2026-06-11"
format:
  revealjs:
    theme: [default, ../../assets/custom.scss]
    slide-number: true
    show-slide-number: all
    progress: true
    history: false
    hash: true
    center: false
    transition: fade
    transition-speed: fast
    code-line-numbers: true
    highlight-style: github
    fig-align: center
    fig-cap-location: bottom
    width: 1280
    height: 720
    margin: 0.1
    self-contained: true
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
#| eval: true
library(tidyverse)
library(fixest)
library(broom)
library(patchwork)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(scales)
library(gapminder)
```

## Today's session {.center}

::: columns
::: {.column width="50%"}
**Part 1 — AI tools**

- What AI is (and isn't) doing
- Three modes of use
- Three failure modes
- The invariant habit

*Live demo: seeing failure in real time*
:::

::: {.column width="50%" .fragment}
**Part 2 — Publication-ready figures**

- Avoid ugly, bad, or wrong graphs
- Choosing the right chart
- Five key extensions
- The 7-item checklist

*Live demo + exercise*
:::
:::

. . .

::: {.callout-note}
The tools will change every six months. The habits we build today will not.
:::

---

# Part 1
## What AI is (and isn't) doing

AI **predicts plausible next tokens** given its training data.

. . .

It does **not**:

- reason about your problem
- look things up or verify claims
- know what has changed since its training cutoff

. . .

**The consequence:**

> Confidence is not correctness.
> Fluent-looking code is not correct code.

## What AI is (and isn't) doing

Hallucination is **structural**, not a glitch to be patched away.

. . .

::: {.callout-tip}
A model trained on Stack Overflow posts from 2022 will confidently emit syntax that no longer works in ggplot2 3.4.0.
:::

. . .

::: {.callout-note}
**What about AI with web search?**
Tools like Perplexity or Claude with research mode retrieve current text — partially lifting the training-cutoff limitation. But retrieved text is still *summarised by a model that cannot verify it*. The hallucination and reasoning limitations remain.
:::

---

## Three modes of use

:::: columns
::: {.column width="33%" .fragment}
### Autocomplete
::: {.incremental}
- AI completes boilerplate you already know how to write

- **Risk:** low
- **Gain:** modest

- Use when you would have typed it anyway
:::
:::

::: {.column width="33%" .fragment}
### Thought partner
::: {.incremental}
- Ask for options, strategies, chart recommendations — **you decide**

- **Risk:** medium
- **Gain:** high

- Works best when you have strong domain knowledge
:::
:::

::: {.column width="33%" .fragment}
### Code reviewer
::: {.incremental}
- Paste your own code; ask what is wrong or what a referee would criticise

- **Risk:** low
- **Gain:** builds critical skill

- Extremely useful, low-risk
:::
:::
::::

. . .

::: {.callout-important}
**The choice of mode is itself the skill.** The session returns to this after every demo.
:::

. . .

::: {.callout-note}
**Modes combine.** A typical workflow: use *thought partner* to sketch a plan, correct it yourself, then *autocomplete* fills in the code. Use *code reviewer* afterwards to catch what you missed.
:::

---

## Three failure modes

::: incremental
1. **Package hallucination**
   AI invents plausible-sounding packages or functions that do not exist.
   A USENIX Security 2025 study across 16 models found **~20% of recommended packages were hallucinations**.
   In R, this is silent until `library()` throws an error.

2. **Stale API knowledge**
   Deprecated arguments (`size=` → `linewidth=` in ggplot2 3.4.0), renamed functions, moved packages.
   The model was trained on older code. It cannot know what changed.

3. **Vibe coding**
   Accepting AI output without reading it. Re-prompting when it breaks without diagnosing why.
   Each unread iteration makes the next harder to understand.
:::

---

## The invariant habit

```
prompt  →  inspect  →  run  →  diagnose
   ↑                              │
   └──────────────────────────────┘
```

. . .

**Inspect** — read the code *before* running it. Does it do what you asked?

**Diagnose** — when it breaks, understand *why* before re-prompting.

. . .

::: {.callout-note}
Students who skip *inspect* and *diagnose* are vibe coding, even if they do not call it that.

This habit is what makes AI use durable as tools change. Everything else in this session may be obsolete in two years. This will not be.
:::

---

## Live demo {.center}

::: {.callout-tip}
**Watch for:**

- Which mode is being used at each step?
- When does AI produce something plausible but wrong?
- What does the *diagnose* step look like in practice?
:::

---

# Part 2

## Two figures: same data

::: columns
::: {.column width="50%"}
**Exploratory** — default output

```{r}
#| label: fig-exploratory
#| echo: false
#| eval: true
#| fig-width: 5
#| fig-height: 3.8
ggplot(gapminder,
       aes(x = gdpPercap,
           y = lifeExp)) +
  geom_point()
```
:::

::: {.column width="50%"}
**Publication-ready**

```{r}
#| label: fig-pubready
#| echo: false
#| eval: true
#| fig-width: 5
#| fig-height: 3.8
gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = gdpPercap,
             y = lifeExp,
             color = continent)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_text_repel(
    aes(label = country),
    data = \(d) filter(d, country == "Germany"),
    size = 4, nudge_x = -1.5, nudge_y = -2.0,
    min.segment.length = 0, 
    show.legend = FALSE
  ) +
  geom_text_repel(
    aes(label = country),
    data = \(d) filter(d, country == "Afghanistan"),
    size = 4, nudge_x = -0.75, nudge_y = -3,
    min.segment.length = 0, 
    show.legend = FALSE
  ) +
  geom_text_repel(
    aes(label = country),
    data = \(d) filter(d, country == "China"),
    size = 4, nudge_x = 0.5, nudge_y = -10,
    min.segment.length = 0, 
    show.legend = FALSE
  ) +
  scale_x_log10(
    labels = label_dollar(suffix = "k", scale = 0.001)) +
  scale_color_viridis_d() +
  theme_classic(base_size = 11) +
  labs(
    x = "GDP per capita (USD, log scale)",
    y = "Life expectancy (years)",
    color = NULL,
    caption = "Source: Gapminder. Year: 2007."
  )
```
:::
:::


---

## Two figures: the code

::: columns
::: {.column width="50%"}
**Exploratory**

```{r}
#| eval: false
ggplot(gapminder,
       aes(x = gdpPercap,
           y = lifeExp)) +
  geom_point()
```

*5 lines. Default output.*
:::

::: {.column width="50%"}
**Publication-ready**

```{r}
#| eval: false
gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = gdpPercap,
             y = lifeExp,
             color = continent)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_text_repel(
    aes(label = country),
    data = \(d) filter(d, country == "Germany"),
    size = 4, nudge_x = -1.5, nudge_y = -2.0,
    min.segment.length = 0, 
    show.legend = FALSE
  ) +
  geom_text_repel(
    aes(label = country),
    data = \(d) filter(d, country == "Afghanistan"),
    size = 4, nudge_x = -0.75, nudge_y = -3,
    min.segment.length = 0, 
    show.legend = FALSE
  ) +
  geom_text_repel(
    aes(label = country),
    data = \(d) filter(d, country == "China"),
    size = 4, nudge_x = 0.5, nudge_y = -10,
    min.segment.length = 0, 
    show.legend = FALSE
  ) +
  scale_x_log10(
    labels = label_dollar(suffix = "k", scale = 0.001)) +
  scale_color_viridis_d() +
  theme_classic(base_size = 11) +
  labs(
    x = "GDP per capita (USD, log scale)",
    y = "Life expectancy (years)",
    color = NULL,
    caption = "Source: Gapminder. Year: 2007."
  )
```
:::
:::

---

## Ugly, bad, wrong

Three different problems — three different fixes.

| Label | Problem | Example | Fix |
|---|---|---|---|
| **Ugly** | Aesthetic | Grey background, `gdpPercap` as axis label | `theme_classic()`, `labs()` |
| **Bad** | Perceptual | Wrong chart type, overlapping labels, spaghetti | Rethink the encoding |
| **Wrong** | Mathematical | No uncertainty, truncated axis, wrong aggregation | Fix the underlying logic |

. . .

::: {.callout-important}
"It doesn't look great" is not a diagnosis. $\rightarrow$ 
"The y-axis is truncated and the comparison is therefore wrong" is.
:::

---

## Regression results — Coefficient plot

*"What predicts life expectancy, and how uncertain are we?"*

```{r}
#| label: fig-coefplot
#| echo: false
#| eval: true
#| fig-width: 8
#| fig-height: 4
gap07 <- gapminder |> filter(year == 2007)
fit   <- lm(lifeExp ~ log(gdpPercap) + continent, data = gap07)
tidy(fit, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(x = estimate,
             y = reorder(term, estimate),
             xmin = conf.low,
             xmax = conf.high)) +
  geom_pointrange(colour = "#1e4a7b") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  theme_classic(base_size = 11) +
  labs(x = "Coefficient estimate (95% CI)", y = NULL,
       caption = "OLS. Source: Gapminder 2007.")
```

::: {.callout-note}
Key geom: `geom_pointrange()` — estimate + uncertainty in one layer
:::

---

## Distributions across groups — Violin + jitter

*"How does life expectancy distribute across continents?"*

```{r}
#| label: fig-violin
#| echo: false
#| eval: true
#| fig-width: 8
#| fig-height: 4
gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = reorder(continent, lifeExp, median),
             y = lifeExp, fill = continent)) +
  geom_violin(alpha = 0.6, linewidth = 0.3, trim = FALSE) +
  geom_jitter(width = 0.07, alpha = 0.5, size = 1.5, colour = "grey30") +
  scale_fill_viridis_d(guide = "none") +
  coord_flip() +
  theme_classic(base_size = 11) +
  labs(x = NULL, y = "Life expectancy (years)",
       caption = "Source: Gapminder 2007.")
```

::: {.callout-note}
Key geoms: `geom_violin()` + `geom_jitter()` — shape AND raw data
:::

---

## Bivariate with heterogeneity — Faceted scatter

*"Does the income–health relationship vary by continent?"*

```{r}
#| label: fig-faceted
#| echo: false
#| eval: true
#| fig-width: 9
#| fig-height: 4
gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(alpha = 0.5, size = 1.5, colour = "#1e4a7b") +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, colour = "#2d8a4e") +
  scale_x_log10(labels = label_dollar(scale = 0.001, suffix = "k")) +
  facet_wrap(~continent, nrow = 2) +
  theme_classic(base_size = 10) +
  labs(x = "GDP per capita (USD, log scale)", y = "Life expectancy (years)",
       caption = "Source: Gapminder 2007.")
```

::: {.callout-note}
Key geoms: `geom_smooth(method = "lm")` + `facet_wrap()` — relationship × group
:::

---

## Time series, multiple units — Highlighted lines

*"How has life expectancy evolved across European countries?"*

```{r}
#| label: fig-lines
#| echo: false
#| eval: true
#| fig-width: 8
#| fig-height: 4
gapminder |>
  filter(continent == "Europe") |>
  ggplot(aes(x = year, y = lifeExp, group = country)) +
  geom_line(linewidth = 0.6) +
  gghighlight(country %in% c("Germany", "Bulgaria"),
              use_direct_label = TRUE,
              label_params = list(size = 3.5)) +
  theme_classic(base_size = 11) +
  labs(x = NULL, y = "Life expectancy (years)",
       caption = "Source: Gapminder.")
```

::: {.callout-note}
Key geoms: `geom_line()` + `gghighlight()` — focus without removing context
:::

---

## Event study / DiD — Centred point-range

*"Did the policy have an effect, and when did it start?"*

```{r}
#| label: fig-eventstudy
#| echo: false
#| eval: true
#| fig-width: 8
#| fig-height: 4
tibble(
  period    = -4:5,
  estimate  = c(-0.1, 0.05, -0.2, 0.1, 0.0, 0.8, 1.4, 1.9, 2.1, 2.3),
  conf_low  = estimate - c(0.40, 0.35, 0.40, 0.35, 0.0, 0.35, 0.38, 0.40, 0.42, 0.45),
  conf_high = estimate + c(0.40, 0.35, 0.40, 0.35, 0.0, 0.35, 0.38, 0.40, 0.42, 0.45)
) |>
  ggplot(aes(x = period, y = estimate,
             ymin = conf_low, ymax = conf_high)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_vline(xintercept = -0.5, linetype = "dashed", colour = "#1e4a7b") +
  geom_pointrange(colour = "#1e4a7b") +
  theme_classic(base_size = 11) +
  labs(x = "Periods relative to treatment",
       y = "Effect estimate (95% CI)",
       caption = "Simulated event study data.")
```

::: {.callout-note}
Key geoms: `geom_pointrange()` + `geom_vline()` — pre/post symmetry + event marker
:::

---

## Choosing the right chart

::: {.callout-warning}
AI defaults to whatever was most common in its training data — usually a bar chart.

**You must choose the chart type before you prompt.**
:::

| Question type | Chart | Key geom(s) |
|---|---|---|
| Regression coefficients | Coefficient plot | `geom_pointrange()` |
| Distributions across groups | Violin + jitter | `geom_violin()` + `geom_jitter()` |
| Bivariate with heterogeneity | Faceted scatter | `geom_smooth()` + `facet_wrap()` |
| Time series, multiple units | Highlighted lines | `geom_line()` + `gghighlight()` |
| Event study / DiD | Centered point-range | `geom_pointrange()` + `geom_vline()` |

---

## The 7-item publication checklist

::: columns
::: {.column width="55%"}
:::{.incremental}
1. Axis labels with **units**; no raw variable names
2. **Colorblind-safe** palette that fits your data (e.g., viridis or Okabe-Ito)
3. **Uncertainty** on every estimate
4. `theme_classic()` (or similar) — no default grey background
5. Base size ~10–12pt for a 3.5-inch figure
6. **Source/notes** in `labs(caption = ...)`
7. `ggsave()` with explicit dimensions — **never** the RStudio Export button
:::
:::

::: {.column width="45%" .fragment}
```{r}
#| eval: false
ggsave(
  "figures/fig1.pdf",
  plot = p,
  width  = 5.5,
  height = 3.5,
  units  = "in",
  device = cairo_pdf
)
```

::: {.callout-tip}
`cairo_pdf` embeds fonts correctly for journal submission.
:::
:::
:::

---

## Five extensions worth knowing

::: columns
::: {.column width="50%"}
**`patchwork`** — multi-panel layout

```{r}
#| eval: false
(p1 | p2) / p3 +
  plot_annotation(
    tag_levels = "A"
  ) +
  plot_layout(
    guides = "collect"
  )
```

**`ggrepel`** — non-overlapping labels

```{r}
#| eval: false
geom_text_repel(aes(label = country))
```

**`gghighlight`** — focus attention

```{r}
#| eval: false
gghighlight(continent == "Europe")
```
:::

::: {.column width="50%"}
**`ggtext`** — Markdown in titles

```{r}
#| eval: false
labs(
  title = "**Treated** firms earn more"
) +
theme(
  plot.title = element_markdown()
)
```

**`scales`** — axis formatting

```{r}
#| eval: false
scale_y_continuous(
  labels = label_comma()
)
scale_x_log10(
  labels = label_dollar()
)
```
:::
:::

---

## Where AI helps — and where it does not

::: columns
::: {.column width="50%" .fragment}
**AI is strong on:**

:::{.incremental}
- Scaffolding a first draft quickly
- Explaining error messages
- Adjusting individual aesthetic parameters
- Refactoring code to use the pipe
:::
:::

::: {.column width="50%" .fragment}
**AI struggles with:**

:::{.incremental}
- Choosing the right chart for the question
- Multi-panel composition and shared legends (`patchwork`)
- Up-to-date ggplot2 API (`linewidth` vs `size`)
- Journal-specific constraints it was not told about
- Self-correcting when a hallucinated package fails
:::
:::
:::

. . .

::: {.callout-note}
AI can get you to about 70% of a publication-ready figure. The last 30% requires judgment you cannot delegate.
:::

---

## Live demo {.center}

::: {.callout-tip}
**Watch for:**

- When and how do we choose the chart type?
- Where does AI succeed at scaffolding?
- Where do we need to step in as the human reviewer?
:::

---

## Exercise {.center}

Available via Github Classroom

::: columns
::: {.column width="50%" .fragment}
**Track A** — suggested starting points

- WMS: distribution of management scores across countries
- CPS: wage–education relationship by subgroup
:::

::: {.column width="50%" .fragment}
**Track B** — bring your own question

- State it in one sentence
- If you cannot, use Track A
:::
:::

. . .

::: {.callout-important}
**Before opening any AI tool:**
Write one sentence justifying your chart type choice.
:::

---

## Quarto + AI: three patterns

**Error diagnosis** — paste a render error with the surrounding chunk

> *"What is causing this error?"*

Reliable on YAML errors and chunk option problems.

. . .

**Goal-finding** — ask how to achieve a specific Quarto output

> *"How do I make a figure span the full slide width in Reveal.js?"*

AI as documentation shortcut — always verify at `quarto.org`.

. . .

**Prose refinement** — describe your figure and ask whether the claim holds

> *"Is this claim supported by the figure? Can it be more concise?"*

Most tools accept a screenshot — share one directly. If yours does not, describing the figure forces the clarity the caption needs anyway.

---

## Closing debrief {.center}

::: columns
::: {.column width="33%"}
**One thing AI is genuinely useful for** in your workflow
:::

::: {.column width="33%"}
**One thing you would never fully delegate** to AI
:::

::: {.column width="33%"}
**One habit** from today you will keep using after the course
:::
:::

. . .

::: {.callout-note}
**Next session:** Recap — send me your question until the end of the week.
:::
