---
title: "Report Title"
subtitle: "Data Science and Econometrics for International Business"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true        # show/hide code blocks with a toggle
    code-tools: true       # copy-code button
  docx:
    toc: true
    number-sections: true
    reference-doc: null    # add a .docx reference file here for custom Word styles
bibliography: ../templates/references.bib   # adjust path as needed
---

```{r}
#| label: setup
#| include: false

library(tidyverse)

# Set a consistent ggplot theme for all figures
theme_set(theme_minimal(base_size = 12))
```

## Introduction

*State your research question here. Lead with the business or policy problem, not the method.*

This report analyses [topic] using data from [source]. The central question is: **[question]?**

## Data

*Briefly describe the dataset: source, unit of observation, key variables, sample period.*

```{r}
#| label: load-data

# data <- read_csv("../data/your-dataset.csv")
```

## Results

*Present your main findings. Interpret coefficients in plain language — avoid reporting numbers without explanation.*

```{r}
#| label: fig-main
#| fig-cap: "Caption describing what the figure shows."

# Your ggplot2 code here
```

As shown in @fig-main, ...

## Discussion

*What do the results mean for the business question? What are the limitations of your analysis?*

## Conclusion

*Summarise the key finding in 2–3 sentences for a non-technical reader.*

## References
