---
title: "Multiple Regression: Going Beyond the Basics"
subtitle: "Session 2 — 09 April 2026"
author: "Prof. Dr. Claudius Gräbner-Radkowitsch"
institute: "Europa-Universität Flensburg"
date: "09 April 2026"
format:
  revealjs:
    theme: [default, ../../assets/custom.scss]
    slide-number: true
    progress: true
    embed-resources: true
    footer: "Adv. Data Science & Econometrics 2026 | Session 2"
    width: 1280
    height: 720
execute:
  eval: false
---

```{r}
#| label: data-setup
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(tidyverse)
library(modelsummary)
library(broom)
library(flextable)
library(kableExtra)

cps_raw <- read_csv("data/morg-2014-emp.csv")

cps <- cps_raw |>
  filter(occ2012 == 735, age >= 24, age <= 64, uhours >= 20) |>
  mutate(
    college = ifelse(grade92 >= 43, 1, 0),
    female  = ifelse(sex == 2, 1, 0),
    age_sq  = age^2,
    edu_cat = case_when(
      grade92 <  43 ~ "No degree",
      grade92 <= 44 ~ "Bachelor's",
      grade92 >= 45 ~ "Graduate"
    ),
    edu_cat = factor(edu_cat, levels = c("No degree", "Bachelor's", "Graduate"))
  ) |>
  select(earnwke, age, age_sq, uhours, college, female, edu_cat) |>
  drop_na()

m1 <- lm(earnwke ~ age,                             data = cps)
m2 <- lm(earnwke ~ age + age_sq,                    data = cps)
m3 <- lm(earnwke ~ age + age_sq + uhours + college, data = cps)
m4 <- lm(earnwke ~ age + age_sq + uhours + college + female, data = cps)
```

## Today's plan

<br>

| Time | Activity |
|---|---|
| 16:00–16:10 | Debriefing of take-home task 1 |
| 16:10–16:30 | Input lecture: categorical variables and interactions |
| 16:30–17:00 | Live coding: expanding take-home 1 |
| 17:00–17:15 | Break |
| 17:15–18:15 | In-session exercise: hotel prices in Vienna |
| 18:15–18:20 | Short break |
| 18:20–18:45 | Debriefing the exercise |
| 18:45–19:00 | Introduce Take-home Task 2, Q&A |

# Debriefing Take Home 1

## Take-home task 1: what you did

**Dataset:** CPS 2014 — market research analysts and marketing specialists

**What you built:**

- Model 1: earnings ~ age (linear)
- Model 2: earnings ~ age + age² (quadratic — better residual pattern)
- Model 3: earnings ~ age + age² + hours + college (with controls)


---

## Take-home task 1: the results {.smaller}

<br>

```{r}
#| echo: false
#| eval: true

modelsummary(
  list("M1: Age" = m1, "M2: Quadratic" = m2, "M3: Full model" = m3),
  coef_map = c("age"         = "Age",
               "age_sq"      = "Age²",
               "uhours"      = "Weekly hours",
               "college"     = "College degree",
               "(Intercept)" = "Constant"),
  gof_map  = c("nobs", "r.squared"),
  stars    = TRUE,
  notes    = "CPS 2014. Market research analysts, age 24–64, ≥ 20 hrs/week."
)
```


. . .

**Be aware:** Adding `uhours` and `college` shifted the age coefficients!

---


## Common issues: interpretation

**Too vague:**
*"The college coefficient is 128.4."*

. . .

**Too mechanical:**
*"When college equals one, earnings increase by 128.4, holding other variables constant."*

. . .

**Good:**
*"Analysts with at least a college degree earn, on average, approximately $128 more per week than those without, holding age and weekly hours constant."*

. . .

::: {.callout-tip}
Always answer: compared to *whom*? By *how much*? In *what units*? Holding *what* constant?
:::

---

## Two new tools for Session 2

**From Task 1 onwards you explored results using raw R output.**

Session 2 introduces two tools that turn raw output into **presentation-ready tables:**

. . .

| Tool | Purpose |
|---|---|
| `modelsummary` | Regression tables — one function for all models |
| `flextable` | Data summaries — any tibble, cleanly formatted |

. . .

Both render in HTML and PDF without any manual formatting.
You will use both in **Take-home Task 2**.

---

## Raw output using `summary()`

```{r}
#| echo: true
#| eval: true

summary(m1)
```

. . .

Not nice, not straightforward to compare several models

## `modelsummary`: before {.smaller}

```{r}
#| echo: true
#| eval: true

modelsummary(list(m1, m2, m3), gof_map = c("nobs", "r.squared"))
```

Column headers auto-named; row labels are raw R names: `age_sq`, `(Intercept)`, `college`

---

## `modelsummary`: after {.smaller}

```{r}
#| echo: true
#| eval: true
#| code-line-numbers: "2|3-5|6|7"

modelsummary(
  list("M1: Age" = m1, "M2: Quadratic" = m2, "M3: Full model" = m3),
  coef_map = c(
    "age"="Age",  "age_sq"="Age²", "uhours" = "Weekly hours", 
    "college" = "College degree", "(Intercept)" = "Constant"),
  gof_map  = c("nobs", "r.squared"),
  notes    = "CPS 2014. Market research analysts, age 24–64."
)
```

---

## `flextable`: before {.smaller}

**Without `flextable` — raw tibble printed to the console:**

```{r}
#| echo: true
#| eval: true

cps |>
  group_by(edu_cat) |>
  summarise(mean_earn = mean(earnwke))
```

→ R's default formatting: variable names as headers, too many decimals, no units.

---

## `flextable`: after {.smaller}

```{r}
#| echo: true
#| eval: true

cps |>
  group_by(edu_cat) |>
  summarise(`Mean weekly earnings (USD)` = round(mean(earnwke), 1)) |>
  rename(`Education level` = edu_cat) |>
  flextable() |>
  theme_vanilla() |>
  autofit()
```

::: {.incremental}
- **`rename()`** — readable column headers before passing to flextable
- **`round()`** — control decimal places in the summary step
- **`theme_vanilla()`** — clean, publication-style formatting
- **`autofit()`** — adjusts column widths to content automatically
:::

---

## The answer to Reflection Question 1

*"Your analysis left out one important individual characteristic that is systematically linked to earnings. Which variable is this?"*

. . .

**Gender.**

. . . 

```{r}
#| echo: true
#| eval: true

tidy(cor.test(cps$female, cps$earnwke)) |>
  select(estimate, p.value, conf.low, conf.high) |>
  mutate(across(everything(), \(x) round(x, 3))) |> 
  flextable() |> 
  theme_vanilla()
```

. . .

Today we learn more about how to properly include categorial variables such as gender

# Categorial variables: introduction

## What is a categorical variable?

A variable that takes a **finite set of named values** — not a continuous scale.

| Variable | Values |
|---|---|
| Gender | Male, Female |
| Education | No degree, Bachelor's, Graduate |
| Hotel star rating | 3-star, 4-star, 5-star |
| Country | Germany, France, Denmark, … |

. . .

**The problem:** You cannot pass the text label `"Female"` to `lm()`.

. . .

**The solution:** Dummy variables — convert each category to a 0/1 indicator.

---

## How R handles categories

R picks one category as the **reference group** and creates one binary (0/1) indicator for each remaining category.

**Example:** Gender (reference = Male)

| Person | Sex | `female` |
|--------|-----|---------------|
| Anna | Female | 1 |
| Ben | Male | 0 |
| Claudia | Female | 1 |

. . .

```r
cps <- cps |> mutate(female = ifelse(sex == 2, 1, 0))
# OR: use factor(sex) and let R create the dummies automatically
```

. . . 

::: {.callout-tip}
Using `factor()` with the `levels` argument lets you set the reference group manually.

The first element in the levels vector will be the reference group (otherwise: defaults to alphabetical order).
:::

---

## Interpreting a dummy coefficient

**Model:** `earnwke ~ age + uhours + female`

. . .

```{r}
#| echo: false
#| eval: true

tidy(lm(earnwke ~ age + uhours + female, data = cps)) |>
  select(term, estimate, std.error, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable(col.names = c("Term", "Estimate", "Std. Error", "p-value"))
```

. . .

→ Female analysts earn, on average, approximately **$`r round(abs(coef(lm(earnwke ~ age + uhours + female, data = cps))["female"]))` per week less** than male analysts of the same age and hours.

---

## Where do dummy coefficients come from?

In a **simple** regression of earnings on gender only, the coefficient equals the raw mean difference:

. . . 

```{r}
#| echo: true
#| eval: true
#| code-line-numbers: "2|3|4"

cps |>
  group_by(female) |>
  summarise(mean_earn = round(mean(earnwke), 1)) |>
  mutate(diff_vs_male = round(mean_earn - mean_earn[female == 0], 1)) |> 
  flextable() |> 
  theme_vanilla()
```

. . . 

```{r}
#| echo: true
#| eval: true

coef(lm(earnwke ~ female, data = cps))["female"] |> round(1)
```

. . .

::: {.callout-important}
With controls (age, uhours), the coefficient changes — it is the difference **holding other variables constant**, not the raw group gap.
:::

---

## Multi-category variables

What if we want education as three levels, not a binary college/no-college split?

::: {.incremental}
- Pick a **reference category** (e.g. "No degree")
- R creates k − 1 dummy variables for k categories
- Each coefficient = difference from the reference group
:::

. . .

```{r}
#| echo: true
#| eval: true

cps |>
  group_by(edu_cat) |>
  summarise(mean_earn = round(mean(earnwke), 1)) |>
  mutate(vs_no_degree = round(mean_earn - mean_earn[edu_cat == "No degree"], 1)) |>
  flextable() |>
  theme_vanilla() |> 
  autofit()
```

---

## Why the reference category matters

**Same model, different reference — different numbers, same underlying story.**

::: {.incremental}
- Reference = No degree: Bachelor +$80, Graduate +$180
- Reference = Bachelor: No degree −$80, Graduate +$100
- Reference = Graduate: No degree −$180, Bachelor −$100
:::

. . .

**How does R choose the reference by default?**

R orders factor levels **alphabetically** — the first level alphabetically becomes the reference.

```r
factor(c("Graduate", "No degree", "Bachelor"))
# Levels: Bachelor's  Graduate  No degree   ← Bachelor is reference (B < G < N)
```

. . . 

**Easiest fix:** set levels manually

```r
factor(c("Graduate", "No degree", "Bachelor's"), levels = c("No degree", "Bachelor", "Graduate"))
# Levels: Bachelor's  Graduate  No degree   ← No degree is reference (N < B < G)
```

. . .

**Ex-post override:** specify the order explicitly or use `relevel()`:

```r
factor(edu, levels = c("No degree", "Bachelor's", "Graduate"))
# Now "No degree" is the reference
relevel(factor(edu), ref = "No degree")  # same result
```

# Interaction terms

## Interaction terms: motivation

**A regression without interactions assumes:**
The effect of age on earnings is *the same* for men and women.

. . .

**But is that realistic?**

- Career interruptions are not equally distributed across genders
- Seniority and promotion patterns differ
- The age–earnings profile may be steeper for men

. . .

**An interaction term lets the slope of age differ by gender.**

---

## Visualising an interaction {.smaller}

```{r}
#| echo: true
#| eval: true
#| fig-height: 4
#| message: false

cps |>
  mutate(gender = ifelse(female == 1, "Female", "Male")) |>
  ggplot(aes(x = age, y = earnwke, colour = gender)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_colour_manual(values = c("Male" = "#00395B", "Female" = "#69AACD")) +
  labs(x = "Age", y = "Weekly earnings (USD)", colour = NULL,
       title = "Are the age–earnings slopes the same for men and women?") +
  theme_minimal()
```

. . . 

But do the slopes really differ significantly, or is this really just a level difference plus noise? 🤔

---

## Interpreting an interaction coefficient {.smaller}

**Model:** `earnwke ~ age * female`  *(shorthand for: `age + female + age:female`)*

```{r}
#| echo: false
#| eval: true

tidy(lm(earnwke ~ age * female, data = cps)) |>
  select(term, estimate, std.error, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(col.names = c("Term", "Estimate", "Std. Error", "p-value"))
```

. . .

::: {.incremental}
- `age`: slope of age **for men** — each additional year adds $[age] to male earnings
- `female`: level difference at age = 0 — not directly interpretable
- `age:female`: *extra* slope for women — how much the age slope **differs** for women
:::

---

## The marginal effect of age

With interactions, "the effect of age" depends on the group.

```{r}
#| echo: false
#| eval: true

tidy(lm(earnwke ~ age * female, data = cps)) |>
  select(term, estimate, std.error, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(col.names = c("Term", "Estimate", "Std. Error", "p-value")) |>
  kableExtra::kable_styling(font_size = 18)
```

```{r}
#| echo: true
#| eval: true

b <- coef(lm(earnwke ~ age * female, data = cps))
```

::: {.incremental}
- Age slope for men: `round(b["age"], 2)` = `r round(b["age"], 2)` 
- Age slope for women: `round(b["age"] + b["age:female"], 2)` = `r round(b["age"] + b["age:female"], 2)`
- Difference: `round(b["age:female"], 2)` = `r round(b["age:female"], 2)`
:::


. . .

$$\frac{\partial \text{earnwke}}{\partial \text{age}} = \hat{\beta}_\text{age} + \hat{\beta}_{\text{age:female}} \times \text{female}$$

→ Always report group-specific slopes — the interaction coefficient alone is difficult to communicate.


# Live coding demo

## Live coding demo

We return to the CPS data and extend the take-home analysis:

1. Add `female` back from the raw CPS file
2. Fit Model 3 from the take-home, now including gender
3. Recode `grade92` from binary to three education categories
4. Visualise: do male and female age-earnings profiles look parallel?
5. Add an age × female interaction
6. Compare all models in one clean `modelsummary` table

::: {.callout-note}
The full code and output are available as **lecture notes** on the course website.
You do not need to take notes — focus on understanding the steps.
:::

# Exercise

## Exercise: hotel prices in Vienna

**Business question:**
*What drives hotel prices — and does the price penalty for being far from the city centre differ between budget and luxury hotels?*

**Dataset:** 428 hotels in Vienna (Békés & Kézdi 2021)
Download from OSF: [osf.io/4e6d8](https://osf.io/4e6d8/)

**Your tasks:**

::: {.incremental}
1. Explore the data — understand what one observation represents
2. Model price as a function of distance and stars (numeric, then as a factor)
3. Add an interaction: does the distance effect vary by star category?
4. Present all models in a clean `modelsummary` table
5. Write a plain-language paragraph for a non-technical manager
:::

# Take-Home Task 2

## Take-Home Task 2

**Assigned today — due 30 April 2026, 08:00**

**Dataset:** Hotels Europe — prices and features across 46 European cities (Békés & Kézdi 2021)

::: {.incremental}
- Build four regression models of increasing complexity (distance → city type → stars → interaction)
- Use `modelsummary` for regression tables and `flextable` for descriptive statistics
- Visualise the price–distance relationship before interpreting the model
- Reflect critically on what your model leaves out
:::

. . .

Full task description distributed via **GitHub Classroom** (link on the course website).

---

## See you on 30 April

**Next session:** Thu, 30 April — *Modelling binary and categorical outcomes*

**Questions during the break?**
[GitHub Discussions](https://github.com/data-science-ma/AdvancedDataScience26-Material/discussions)

**Task 2 deadline:** 29 April 2026, 23:59

. . .

The lecture notes extending Task 1 are available now on the course website.

::: aside
**PDF export (Chrome/Edge):** click ☰ bottom-left → Tools → PDF Export Mode → print.
:::
