---
title: "What Can Go Wrong"
subtitle: "Biases and Diagnostics · Session 4"
author: "Claudius Gräbner-Radkowitsch"
date: "2026-05-21"
format:
  revealjs:
    theme: [default, ../../assets/custom.scss]
    slide-number: true
    progress: true
    code-line-numbers: false
    highlight-style: github
    fig-align: center
    self-contained: true
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(broom)
library(modelsummary)
library(lmtest)
library(sandwich)
library(car)
library(performance)
library(see)
library(patchwork)

hotels <- read_csv("data/hotels_cs.csv", show_col_types = FALSE)

theme_diag <- theme_minimal(base_size = 14) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    axis.title    = element_text(size = 11)
  )

model_base <- lm(price ~ distance + stars, data = hotels)
model_work <- lm(log(price) ~ log(distance + 0.1) + stars, data = hotels)
```

## Today's session {.center}

::: {.incremental}
1. When can we trust OLS output? — the Gauss-Markov framework
2. Functional form and the linearity assumption
3. Heteroskedasticity and robust standard errors
4. Multicollinearity
5. Influential observations
6. Normality of residuals
7. What diagnostics *cannot* detect: omitted variable bias
:::

---

## A regression output we need to interrogate {.smaller}

We have estimated hotel price as a function of distance and star rating.

::: {.incremental}
- Are these coefficient estimates pointing in the right direction?
- Are the standard errors — and the stars — reliable?
- Is there anything the diagnostics *cannot* reveal?
:::

. . .

```{r}
#| echo: false
modelsummary(model_base, stars = TRUE,
             gof_omit = "AIC|BIC|Log|F|RMSE",
             fmt = 2, output = "markdown")
```

. . . 

- This requires **theoretical considerations** _and_ **formal diagnostics**

# The OLS Contract {.center}

## BLUE: a guarantee with conditions

**B**est **L**inear **U**nbiased **E**stimator — OLS is BLUE *when its assumptions hold*

::: {.incremental}
- Theorem about properties of the OLS **sampling distribution**
- **Unbiased:** on average, the estimate equals the true value — $E[\hat{\beta}] = \beta$
- **Efficient (Best):** among all linear unbiased estimators, OLS has the smallest variance
- These are *different* properties, protected by *different* assumptions
:::

. . .

> Unbiasedness = pointing in the right direction on average
> Efficiency = grouping tightly around the truth

## Unbiasedness and efficiency: visualised

```{r}
#| fig-width: 11
#| fig-height: 4.5
set.seed(999)
B       <- 1500
true_b  <- 2.0
n_full  <- 80

results <- map_dfr(seq_len(B), function(b) {
  x    <- rnorm(n_full)
  z    <- 0.7 * x + rnorm(n_full, 0, sqrt(1 - 0.49))
  y    <- 1 + true_b * x + 1.5 * z + rnorm(n_full)
  xs <- x[1:20]; ys <- y[1:20]; zs <- z[1:20]   # named objects, predictable coef names
  tibble(
    efficient   = coef(lm(y  ~ x  + z))[["x"]],   # correct model, full sample
    biased      = coef(lm(y  ~ x))[["x"]],         # omits z → OVB
    inefficient = coef(lm(ys ~ xs + zs))[["xs"]]   # correct model, small sample → wide
  )
})

label_map <- c(
  efficient   = "OLS — correct model\n(Unbiased + Efficient)",
  biased      = "OLS — omits z\n(Biased)",
  inefficient = "OLS — small sample (n=20)\n(Unbiased, Inefficient)"
)

results |>
  pivot_longer(everything(), names_to = "estimator", values_to = "estimate") |>
  mutate(estimator = factor(label_map[estimator], levels = label_map)) |>
  ggplot(aes(estimate, color = estimator, fill = estimator)) +
  geom_density(alpha = 0.2, linewidth = 0.9) +
  geom_vline(xintercept = true_b, linetype = "dashed",
             color = "gray30", linewidth = 1) +
  annotate("text", x = true_b + 0.1, y = Inf, vjust = 2,
           label = "True β = 2", size = 3.8, color = "gray30") +
  scale_fill_manual(values  = c("steelblue", "tomato", "goldenrod3")) +
  scale_color_manual(values = c("steelblue", "tomato", "goldenrod3")) +
  labs(x = "Estimated β", y = "Density", color = NULL, fill = NULL,
       subtitle = "1,500 simulated datasets — each curve is a sampling distribution") +
  theme_diag +
  theme(legend.position = "bottom", legend.text = element_text(size = 10))
```

*Blue: correct answer, precise. Gold: correct direction, imprecise. Red: wrong answer entirely.*

## The assumption checklist

| Assumption | Formula | Plain English | Cost if violated |
|---|---|---|---|
| **Linearity** | $y = X\beta + u$ | Correct functional form | Biased estimates |
| **No perfect collinearity** | $\text{rank}(X) = k$ | Predictors not identical | Cannot estimate |
| **Exogeneity** | $E[u \mid X] = 0$ | No omitted confounders | Biased estimates |
| **Homoskedasticity** | $\text{Var}(u \mid X) = \sigma^2$ | Constant error variance | SEs are wrong |
| *(Normality)* | $u \mid X \sim \mathcal{N}(0,\sigma^2)$ | *Normal errors* | *Exact inference only* |

. . .

Assumptions 1 & 3 broken → **wrong answer even in large samples**
Assumption 4 broken → **right answer, wrong uncertainty**

## Why $E[u \mid X] = 0$ fails: five routes

::: {.incremental}
- **Omitted variable bias** — a relevant determinant of $y$ is left in $u$ and correlated with $X$ (e.g. ability in wage regressions)
- **Measurement error in regressors** — if $X$ is observed with error correlated with the true regressor, estimates are attenuated or biased
- **Simultaneity / reverse causality** — $X$ affects $y$ *and* $y$ affects $X$; the regressor is jointly determined with the outcome (price–quantity, income–health)
- **Dynamic models with serially correlated errors** — including $y_{t-1}$ as a regressor while $u_t$ is autocorrelated makes the lag endogenous
- **Functional form misspecification** — wrong transformation or omitted interactions push systematic structure into $u$, which then correlates with included regressors
:::

. . .

*All five routes share one consequence: $\hat{\beta}$ is biased and inconsistent — wrong even in large samples.*

## Our diagnostic dashboard

One function covers five of the six checks:

```{r}
#| echo: true
#| eval: false
performance::check_model(model_base)
```

. . .

Each panel targets one assumption — or one practical robustness concern (influential observations).
We will go through them in order.

## Our diagnostic dashboard — output {.center}

```{r}
#| fig-width: 12
#| fig-height: 6.0
#| out-width: "100%"
check_model(model_base)
```

# Functional Form {.center}

## Linearity in parameters ≠ linearity in variables

OLS requires the model to be **linear in the parameters** — not in the raw variables.

. . .

| Transformation | Interpretation of $\hat{\beta}$ |
|---|---|
| `y ~ x` (level–level) | +1 unit of x → $+\hat{\beta}$ units of y |
| `log(y) ~ x` (log–level) | +1 unit of x → $+(\hat{\beta} \times 100)\%$ change in y |
| `y ~ log(x)` (level–log) | +1% change in x → $+\hat{\beta}/100$ units of y |
| `log(y) ~ log(x)` (log–log) | +1% change in x → $+\hat{\beta}\%$ change in y |
| `y ~ x + I(x^2)` (quadratic) | Relationship has a turning point |

. . .

*You already used log transformations in Task 1 — this session extends that logic.*

## Residuals vs. fitted: the primary diagnostic

```{r}
#| fig-width: 11
#| fig-height: 4

set.seed(42)
n <- 300
x <- seq(0.5, 10, length.out = n)

y_lin  <- 1.5 + 1.8 * x + rnorm(n, 0, 1.8)
y_quad <- 8   + 3   * x - 0.35 * x^2 + rnorm(n, 0, 1.2)

m_lin  <- lm(y_lin  ~ x)
m_quad <- lm(y_quad ~ x)

make_resid <- function(m, title, subtitle) {
  augment(m) |>
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = 0.35, color = "steelblue", size = 1.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 1) +
    labs(x = "Fitted values", y = "Residuals",
         title = title, subtitle = subtitle) +
    theme_diag
}

make_resid(m_lin,  "Assumption holds",   "Random scatter — no pattern") +
make_resid(m_quad, "Assumption violated", "Curved pattern — wrong functional form")
```

The red smoother should be **flat**. A systematic curve means something is missing.

## Hotels-europe: linear vs. log-log

```{r}
#| fig-width: 11
#| fig-height: 4

m_linear <- lm(price       ~ distance,            data = hotels)
m_loglog  <- lm(log(price) ~ log(distance + 0.1), data = hotels)

make_resid(m_linear, "Level–level: price ~ distance",         "") +
make_resid(m_loglog, "Log–log: log(price) ~ log(distance)", "")
```

## The RESET test

Formal test for functional form misspecification (`lmtest::resettest()`).

```{r}
#| echo: true
resettest(m_linear)
```

. . .

::: {.callout-warning}
A rejection tells you **something is wrong** — not **what** to change.
Use the residual plot to guide the fix; don't let the test prescribe it.
:::

## Fix 1 — Log transformation for a multiplicative relationship

```{r}
#| echo: true
#| eval: false
lm(log(price) ~ log(distance + 0.1), data = hotels)
```

```{r}
#| fig-width: 11
#| fig-height: 3.8

set.seed(55)
n_m   <- 250
x_m   <- runif(n_m, 0.5, 10)
y_m   <- 3 * x_m^1.5 * exp(rnorm(n_m, 0, 0.4))   # true: power-law / multiplicative

m_m_level <- lm(y_m         ~ x_m)
m_m_log   <- lm(log(y_m)    ~ log(x_m))

make_resid(m_m_level, "Level–level  (misspecified)",
           "Curve reveals multiplicative structure is ignored") +
make_resid(m_m_log,   "Log–log  (corrected)",
           "Residuals are now random — functional form matches the data")
```

. . .

**Interpretation:** $\hat{\beta}$ is an *elasticity* — a 1% increase in $x$ is associated with a $\hat{\beta}$% change in $y$.
*Example:* `log(price) ~ log(distance)` with $\hat{\beta} = -0.12$ → "10% further from centre → price 1.2% lower."

## Fix 2 — Quadratic term for a curved relationship

```{r}
#| echo: true
#| eval: false
lm(price ~ distance + I(distance^2), data = hotels)
```

```{r}
#| fig-width: 11
#| fig-height: 3.8

set.seed(66)
n_q  <- 250
x_q  <- runif(n_q, 0, 10)
y_q  <- 2 + 4 * x_q - 0.38 * x_q^2 + rnorm(n_q, 0, 2)   # true: quadratic

m_q_lin  <- lm(y_q ~ x_q)
m_q_quad <- lm(y_q ~ x_q + I(x_q^2))

make_resid(m_q_lin,  "Linear  (misspecified)",
           "Arch pattern — the relationship has a turning point") +
make_resid(m_q_quad, "Quadratic  (corrected)",
           "Residuals are now random — the turning point is captured")
```

. . .

**Interpretation:** the marginal effect of $x$ is $\hat{\beta}_1 + 2\hat{\beta}_2 x$ — it varies with $x$.
Turning point at $x^* = -\hat{\beta}_1 / (2\hat{\beta}_2)$.
*Note:* with $\hat{\beta}_2 < 0$, the effect of $x$ on $y$ diminishes and eventually reverses.

# Heteroskedasticity {.center}

## Non-constant error variance

OLS assumes $\text{Var}(\varepsilon_i) = \sigma^2$ for all $i$ — the same uncertainty everywhere.

**What goes wrong when this fails:**

::: {.incremental}
- OLS estimates are **still unbiased** — the direction and magnitude are correct on average
- But standard errors are **inconsistent** — too narrow in some places, too wide in others
- t-tests and confidence intervals **mislead** you about precision
:::

. . .

*Example: hotel prices in Istanbul cluster tightly around the regression line; London hotels scatter much more widely. A single SE cannot represent both groups honestly.*

## Scale-location plot: the visual test

```{r}
#| fig-width: 11
#| fig-height: 4

set.seed(101)
n <- 300
x_h <- runif(n, 0.5, 9)

y_homo   <- 1 + 1.5 * x_h + rnorm(n, 0, 2)
y_hetero <- 1 + 1.5 * x_h + rnorm(n, 0, 0.55 * x_h)

m_homo   <- lm(y_homo   ~ x_h)
m_hetero <- lm(y_hetero ~ x_h)

make_scaleloc <- function(m, title, subtitle) {
  augment(m) |>
    ggplot(aes(.fitted, sqrt(abs(.std.resid)))) +
    geom_point(alpha = 0.35, color = "steelblue", size = 1.2) +
    geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 1) +
    labs(x = "Fitted values", y = "√|Standardised residuals|",
         title = title, subtitle = subtitle) +
    theme_diag
}

make_scaleloc(m_homo,   "Assumption holds",   "Flat band — constant spread") +
make_scaleloc(m_hetero, "Assumption violated", "Fan shape — spread increases with fitted values")
```

The red smoother should be **flat and horizontal**.

## Hotels-europe: heteroskedasticity in practice

```{r}
#| fig-width: 10
#| fig-height: 4

augment(model_base) |>
  ggplot(aes(.fitted, sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.2, color = "steelblue", size = 0.9) +
  geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 1.2) +
  labs(x = "Fitted values", y = "√|Standardised residuals|",
       title = "Scale-location plot: hotels-europe",
       subtitle = "price ~ distance + stars  |  November 2017, 8 cities") +
  theme_diag
```

## The Breusch-Pagan test

```{r}
#| echo: true
bptest(model_base)
```

. . .

- **Null:** homoskedasticity (constant variance)
- Rejection → heteroskedasticity present — standard errors are not reliable

. . .

::: {.callout-note}
The test confirms what the plot already showed. Always look at the plot first — the test just formalises the diagnosis.
:::

## The fix: Heteroskedasticity-robust (HC) standard errors {.smaller}

- Standard formula $\rightarrow$ "sandwich" estimator: uses each observation's squared residual to estimate uncertainty.
- **Result:** larger SEs where the model struggles, smaller where it fits well $\rightarrow$ Estimates do not change.

```{r}
#| echo: true
#| eval: false
modelsummary(list("Standard" = model_base, "Robust (HC1)" = model_base),
             vcov = list("classical", "HC1"), stars = TRUE)
```

```{r}
#| echo: false
modelsummary(
  list("Standard" = model_base, "Robust (HC1)" = model_base),
  vcov     = list("classical", "HC1"),
  stars    = TRUE,
  fmt      = 2,
  gof_omit = "AIC|BIC|Log|F|RMSE",
  output   = "markdown"
)
```

. . .

**Rule of thumb:** with cross-sectional data, use robust SEs by default.

# Multicollinearity {.center}

## When predictors move together

**Perfectly correlated** predictors: OLS breaks down

. . .

**Highly correlated** predictors: isolating individual contributions becomes difficult

. . .

**Consequence:**

::: {.incremental}
- **Estimates remain unbiased** — this is an efficiency problem, not a bias problem
- Standard errors **inflate** — imprecise, unstable estimates
- Small changes to the sample can flip signs or change magnitudes substantially
:::

. . .

*Example: imagine adding both `stars` (hotel classification) and `rating` (guest satisfaction score) to the model. Both measure "quality" — how to check whether this is problematic?*

## VIF: the Variance Inflation Factor

**VIF**: by how much is variance of $\hat{\beta}_j$ inflated relative to a model with no correlation among predictors.

$$\text{VIF}_j = \frac{1}{1 - R^2_j}$$

$R^2_j$: the $R^2$ from regressing predictor $j$ on all other predictors.

| VIF | Interpretation |
|---|---|
| 1 | No inflation — predictor is uncorrelated with others |
| 1–5 | Mild — generally acceptable |
| 5–10 | Moderate concern — inspect carefully |
| > 10 | Serious — standard errors substantially inflated |

## Hotels-europe: low vs. problematic VIF

**Base model** — distance and stars measure different things: VIF is fine.

```{r}
#| echo: true
vif(model_base)   # price ~ distance + stars
```

. . .

**Redundant model** — a researcher includes both `stars` and `log(stars)` to "try both functional forms." They are near-identical (r ≈ 0.97): VIF explodes.

```{r}
#| echo: true
model_collinear <- lm(price ~ distance + stars + log(stars), data = hotels)
vif(model_collinear)
```

. . .

::: {.callout-warning}
Including a variable and a monotonic transformation of it simultaneously is the most common source of severe collinearity. The model cannot separate their effects — the SEs become unreliable. Drop one or the other.
:::

## What to do about multicollinearity

::: {.incremental}
- **Do not** drop variables mechanically based on VIF — think about what each variable represents
- Dropping an important control to reduce VIF can introduce omitted variable bias (a far worse problem)
- If two variables genuinely measure the same construct, keep only one — or *combine* them into a single index (e.g. a simple average, or the first principal component)
- Sometimes collinearity is just a feature of reality — acknowledge it and note that individual coefficients should be interpreted with caution
:::

. . .

::: {.callout-warning}
**Interaction terms** (`x * z`) always produce high VIF for their component variables. This is expected and harmless — centering the variables reduces VIF cosmetically but does not change inference.
:::

# Influential Observations {.center}

## Three types of unusual data points

| Type | What makes it unusual | Does it distort results? |
|---|---|---|
| **Outlier** | Large residual — unusual *y* given *x* | Can do, but not always |
| **High-leverage point** | Extreme position in *x* space | Can do, but not always |
| **Influential observation** | Removing it substantially changes the estimates | **Yes, by definition** |

. . .

The dangerous case: a point that is *both* an outlier *and* high-leverage. 

## Cook's distance

$$D_i = \frac{(\hat{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}_{(-i)})^\top (X^\top X) (\hat{\boldsymbol{\beta}} - \hat{\boldsymbol{\beta}}_{(-i)})}{p \cdot \hat{\sigma}^2}$$

. . .

$D_i$: how much do the estimated coefficients shift when observation $i$ is removed.

**Rule of thumb:** $D_i > 4/n$ flags an observation worth investigating.


## Example: one influential point

```{r}
#| fig-width: 11
#| fig-height: 4.5

set.seed(303)
n_base <- 60

df_clean     <- tibble(x = runif(n_base, 0, 6),
                       y = 1 + 1.8 * x + rnorm(n_base, 0, 0.8))
df_influence <- bind_rows(df_clean, tibble(x = 13, y = 3))

m_clean <- lm(y ~ x, data = df_clean)
m_infl  <- lm(y ~ x, data = df_influence)

p_clean <- df_clean |>
  ggplot(aes(x, y)) +
  geom_point(color = "steelblue", alpha = 0.7) +
  geom_abline(slope = coef(m_clean)[2], intercept = coef(m_clean)[1],
              color = "tomato", linewidth = 1.2) +
  coord_cartesian(xlim = c(0, 14), ylim = c(-1, 14)) +
  labs(title = "Without influential point",
       subtitle = sprintf("slope = %.2f", coef(m_clean)[2])) +
  theme_diag

p_infl <- df_influence |>
  mutate(influential = row_number() == n()) |>
  ggplot(aes(x, y, color = influential)) +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c("steelblue", "tomato"), guide = "none") +
  geom_abline(slope = coef(m_clean)[2], intercept = coef(m_clean)[1],
              color = "gray60", linewidth = 0.8, linetype = "dashed") +
  geom_abline(slope = coef(m_infl)[2],  intercept = coef(m_infl)[1],
              color = "tomato", linewidth = 1.2) +
  coord_cartesian(xlim = c(0, 14), ylim = c(-1, 14)) +
  labs(title = "With one influential point (red)",
       subtitle = sprintf("slope = %.2f  (was %.2f)", coef(m_infl)[2], coef(m_clean)[2])) +
  theme_diag

p_clean + p_infl
```

*The dashed line is the original fit. One point pulls the slope down substantially.*

## Hotels-europe: who are the influential hotels?

```{r}
#| echo: true
augment(model_base) |>
  bind_cols(hotels |> select(hotel_id, city)) |>
  arrange(desc(.cooksd)) |>
  select(hotel_id, city, stars, price, .cooksd) |>
  head(8)
```

## Investigate — don't delete

::: {.incremental}
- An influential observation is **not automatically an error**. It may be the most interesting case in the dataset.
- Ask: *why* is this point influential? Data entry error, genuine extreme case, or a different sub-population?
- **Workflow:** re-estimate without the flagged observations. Report *both* results and note the sensitivity.
- Silently removing influential points without disclosure is a **serious methodological error**.
:::

# Normality of Residuals {.center}

## The Q-Q plot: good vs. violated

```{r}
#| fig-width: 11
#| fig-height: 4

set.seed(404)
n    <- 250
x_n  <- runif(n, 0, 10)

y_normal <- 2 + x_n + rnorm(n)
y_skewed <- 2 + x_n + (rexp(n, 0.4) - 2.5)

m_normal <- lm(y_normal ~ x_n)
m_skewed <- lm(y_skewed ~ x_n)

make_qq <- function(m, title, subtitle) {
  augment(m) |>
    ggplot(aes(sample = .std.resid)) +
    stat_qq(alpha = 0.4, color = "steelblue", size = 1.2) +
    stat_qq_line(color = "tomato", linewidth = 1) +
    labs(x = "Theoretical quantiles", y = "Sample quantiles",
         title = title, subtitle = subtitle) +
    theme_diag
}

make_qq(m_normal, "Normal residuals",  "Points follow the diagonal — assumption holds") +
make_qq(m_skewed, "Skewed residuals",  "Systematic departure in the tails — assumption violated")
```

## A widespread misconception

**What normality actually protects:**
Exact finite-sample inference — t- and F-statistics follow their named distributions *exactly* only when errors are normal.

. . .

**Why it rarely matters in practice:**

::: {.incremental}
- With moderate-to-large $n$: the **Central Limit Theorem** ensures $\hat{\beta}$ is approximately normally distributed *regardless of the residual distribution*
- For our hotels dataset ($n \approx 9{,}600$): normality of residuals is essentially irrelevant for inference
- When it does matter: very small samples ($n < 30$), exact prediction intervals
:::

## The Shapiro-Wilk trap

```{r}
#| echo: true
shapiro.test(resid(model_base)[1:5000])  # max n for shapiro.test is 5000
```

. . .

::: {.callout-warning}
With large $n$, Shapiro-Wilk **almost always rejects** — it detects trivially small deviations from normality that have no practical consequence for inference.

**Do not interpret a significant p-value here as a problem.** Check the Q-Q plot visually instead.
:::

# The Posterior Predictive Check {.center}

## What the PPC shows

The PPC simulates new response values from the fitted model many times and compares their distribution to the observed data.



. . .

*The PPC catches what the other five panels cannot: a fundamental mismatch between the model's implied data-generating process and the actual distribution of $y$.*

## PPC: level-level vs. log-log

```{r}
#| fig-width: 12
#| fig-height: 4.5

p_base <- plot(check_predictions(model_base)) +
  labs(title = "Level–level: price ~ distance + stars",
       subtitle = "Residual skewness = 7.8 — simulated prices spread to €3,500+") +
  theme_diag

p_work <- plot(check_predictions(model_work)) +
  labs(title = "Log–log: log(price) ~ log(distance) + stars",
       subtitle = "Residual skewness ≈ 0 — predicted and observed distributions align") +
  theme_diag

p_base + p_work
```

::: {.incremental}
- **Blue line:** density of the *observed* outcome $y$
- **Green lines:** densities of outcomes *simulated* from the fitted model
- If green ≈ blue → the model generates plausible data → distributional assumptions are reasonable
- If they diverge → the model produces values in the wrong range → mis-specified DGP
:::


## Dashboard panels 1–2: functional form and variance

```{r}
#| fig-width: 12
#| fig-height: 4.2
check_model(model_work, check = c("linearity", "homogeneity"))
```

- **Linearity** — residuals vs. fitted; the red smoother should be flat. Systematic curve → functional form problem
- **Homogeneity of variance** — scale-location; the smoother should be horizontal. Remaining fan shape → use robust SEs even with the log model

## Dashboard panels 3–4: collinearity and normality

```{r}
#| fig-width: 12
#| fig-height: 4.2
check_model(model_work, check = c("vif", "qq"))
```

- **Collinearity (VIF)** — bars should be short (< 5). Both predictors are fine; recall the `stars + log(stars)` example where VIF ≈ 20
- **Normality of residuals (Q-Q)** — much closer to the diagonal than `model_base` (skewness −0.1 vs. 7.8); with $n > 9{,}000$ the remaining departure does not affect inference

## Dashboard panels 5–6: influential observations and overall fit

```{r}
#| fig-width: 12
#| fig-height: 4.8
check_model(model_work, check = c("outliers", "pp_check"))
```

- **Outliers / influential observations** — Cook's D bar and leverage–residuals scatter; points outside the contour lines warrant investigation (report with and without, do not silently delete)
- **Posterior predictive check** — green lines now closely track the blue observed density. The log transformation resolved the distributional mismatch we saw with `model_base`.

# What Diagnostics Cannot Detect {.center}

## What residuals cannot tell you

All diagnostic panels show properties of the *residuals* — the part of $y$ the model did not explain.

But every route to endogeneity ($E[u \mid X] \neq 0$) is invisible in the residuals:

::: {.incremental}
- **Omitted variable bias** — a confounder is absorbed into $\hat{\beta}$; the fit looks fine
- **Measurement error in regressors** — attenuation bias hides behind well-behaved residuals
- **Simultaneity / reverse causality** — the endogenous regressor leaves no obvious residual pattern
:::

. . .

All three share the same logic: the bias is *inside* $\hat{\beta}$, not visible in $\hat{u}$.

. . .

We focus on **OVB** — the most common case in business and economics research.

## The mechanism

Suppose the true model is:

$$y = \beta_0 + \beta_1 x + \beta_2 z + \varepsilon$$

but we estimate $y = \beta_0 + \beta_1 x + \varepsilon$ (omitting $z$).

. . .

The estimated $\hat{\beta}_1$ absorbs part of the effect of $z$ — and is biased unless $z$ is *uncorrelated* with $x$.

. . .

**Direction-of-bias rule** (no algebra needed):

$$\text{sign(bias)} = \text{sign}(\text{effect of } z \text{ on } y) \times \text{sign}(\text{corr}(z, x))$$

## OVB in practice: what changes when we add city?

```{r}
#| echo: true
m_no_city   <- lm(log(price) ~ log(distance + 0.1), data = hotels)
m_with_city <- lm(log(price) ~ log(distance + 0.1) + city, data = hotels)
```

```{r}
modelsummary(
  list("Without city" = m_no_city, "With city controls" = m_with_city),
  coef_map = c("log(distance + 0.1)" = "log(distance)"),
  stars    = TRUE,
  fmt      = 3,
  gof_omit = "AIC|BIC|Log|F|RMSE",
  output   = "markdown",
  add_rows = tribble(
    ~term,      ~`Without city`, ~`With city controls`,
    "City FE",  "No",            "Yes"
  )
)
```

. . .

The distance coefficient changes substantially — because **city** drives both price level *and* how far hotels are from the centre.

## Why the residuals looked fine

Run the full diagnostic suite on the *biased* model (no city controls):

```{r}
#| fig-width: 12
#| fig-height: 3.8
check_model(m_no_city, check = c("linearity", "homogeneity"))
```

. . .

Both panels pass $\rightarrow$ residuals scatter randomly, variance looks stable.
**Yet we know from the previous slide that $\hat{\beta}_\text{distance}$ is wrong.**

. . .

::: {.callout-important}
The bias was absorbed into $\hat{\beta}$ when the model was fitted — the residuals have already been "cleaned" of it.
OVB is invisible to any diagnostic tool. The solution is **research design**, not better diagnostics.
:::

## The way forward

**Panel data and fixed effects** (panel data session):

- If the omitted variable is *time-invariant* (e.g. which city), fixed effects eliminate it
- One of the most powerful practical solutions to OVB

. . .

**Causal identification** (optional session on causality):

- Randomisation, quasi-experiments
- Address OVB at the design stage, not the estimation stage

. . .

> The transition from *describing patterns* to *making causal claims* is one of the most important methodological steps in applied data science.

# Summary {.center}

## The diagnostic workflow

For every regression, ask:

::: {.incremental}
1. **Residuals vs. fitted** — is the functional form correct? Is there a pattern?
2. **Scale-location** — is the variance constant? Do I need robust SEs?
3. **Cook's distance / leverage** — are a few observations driving my results?
4. **VIF** — are my predictors too correlated to separate their effects?
5. **Q-Q plot** — are there serious departures from normality? (Large $n$: rarely a concern)
6. **Thinking** — what important variable might I have left out?
:::

. . .

```r
performance::check_model(your_model)  # panels 1–5 in one call
```

The sixth question — OVB — requires subject-matter knowledge, not a function.
