---
title: "Session 6b: Advanced Panel Data"
subtitle: "Live Demo & Lecture Notes"
author: "Claudius Gräbner-Radkowitsch"
date: "Optional — see course website for scheduling"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

## Business question

> *Does ownership type affect management quality — and do firms that
> improve their size also improve their management practices?*

Session 6 established fixed effects as the core panel tool: by
focusing on within-unit variation over time, FE controls for stable
unobserved firm characteristics that would otherwise bias our estimates.

This session covers three situations where FE is not the right choice:

1. **Time-invariant covariates** like ownership type are differenced
   out by FE — we cannot estimate their effects at all
2. **Change questions** are more naturally addressed by first
   differencing consecutive observations rather than deviating from
   unit means
3. **Dynamic outcomes** that depend on their own past values require
   a different estimator entirely — FE is biased in this case

## Packages and data

This session introduces `plm`, the standard R package for RE, FD,
and several other panel estimators that `fixest` does not implement.
We continue to use `fixest` for FE to stay consistent with Session 6.

```{r}
#| label: setup
library(tidyverse)
library(fixest)
library(plm)
library(modelsummary)
```

```{r}
#| label: load-data
wms <- read_csv("data/wms_da_textbook.csv") |>
  mutate(log_emp = log(emp_firm))

# plm requires a pdata.frame that declares the unit and time indices
pdata <- pdata.frame(wms, index = c("firmid", "wave"))

glimpse(wms)
```

::: {.callout-note}
**Why two packages?** `fixest::feols()` is faster and better-designed
for high-dimensional FE — that is why we used it in Session 6 and
continue to use it here for FE models. `plm` covers RE and FD, which
`fixest` does not implement. Always use the right tool for the job.
:::

The key variables are the same as in Session 6. One addition:

| Variable | Description |
|---|---|
| `ownership` | Ownership type: founder, family, dispersed, government, etc. |

`ownership` is largely stable within firms across survey waves — it
is a time-invariant covariate that FE cannot estimate.

---

# Random Effects

## Why FE cannot answer every question

The within-transformation removes *all* time-invariant variation —
including that of covariates we care about substantively.

Let us confirm that `ownership` barely changes within firms:

```{r}
#| label: ownership-variation

wms |>
  group_by(firmid) |>
  summarise(n_ownership_types = n_distinct(ownership, na.rm = TRUE)) |>
  count(n_ownership_types, name = "n_firms")
```

The vast majority of firms have exactly one ownership type across
all waves — confirming that `ownership` is effectively time-invariant
in this dataset. Including it in a FE model will either drop it or
return unreliable estimates.

## The RE model

The Random Effects model treats $\eta_i$ as a random variable drawn
from a distribution with mean zero and variance $\sigma^2_\eta$,
rather than as a fixed unknown parameter per unit. This changes the
estimation from demeaning (FE) to Generalised Least Squares (GLS),
which is a weighted combination of within and between information.

**Consequence:** because unit means are not subtracted, time-invariant
covariates like `ownership` are retained and estimable.

**Cost:** RE is only consistent when $\eta_i$ is uncorrelated with
the regressors. If firms with strong management culture (high $\eta_i$)
also tend to be larger or more educated, RE picks up that correlation
as a spurious covariate effect.

## Estimating FE and RE side by side

`plm()` uses the same formula interface as `lm()`. The `model`
argument selects the estimator:

```{r}
#| label: fe-re-comparison

# FE (within) — our Session 6 baseline
m_fe <- plm(management ~ log_emp + degree_m,
            data  = pdata,
            model = "within")

# RE (GLS)
m_re <- plm(management ~ log_emp + degree_m,
            data  = pdata,
            model = "random")

modelsummary(
  list("FE (within)" = m_fe, "RE" = m_re),
  gof_map = c("nobs", "r.squared"),
  notes   = "FE: within estimator. RE: GLS (Swamy-Arora variance components)."
)
```

The coefficients on `log_emp` and `degree_m` are likely similar here —
both estimators are using much the same variation when $T$ is small.
The real difference appears when we add `ownership`:

```{r}
#| label: re-ownership

# FE with ownership — will silently drop or warn
m_fe_own <- plm(management ~ log_emp + degree_m + ownership,
                data  = pdata,
                model = "within")

# RE with ownership — the quantity FE cannot provide
m_re_own <- plm(management ~ log_emp + degree_m + ownership,
                data  = pdata,
                model = "random")

modelsummary(
  list(
    "FE"           = m_fe,
    "FE + own."    = m_fe_own,
    "RE"           = m_re,
    "RE + own."    = m_re_own
  ),
  gof_map = c("nobs", "r.squared"),
  notes   = "Ownership silently dropped or unreliable in FE models."
)
```

Notice how FE handles the ownership dummies — compare that to the
RE result where all categories are estimated.

## The Hausman test

The critical question is whether the RE assumption is defensible:
does $\eta_i$ correlate with the regressors?

The **Hausman test** answers this by comparing FE and RE slope
estimates for the common covariates (`log_emp`, `degree_m`). Under
$H_0$, both are consistent and the estimates should be similar.
A significant divergence means RE is picking up endogeneity — use FE.

```{r}
#| label: hausman

phtest(m_fe, m_re)
```

**Reading the result:**

- $p < 0.05$: reject $H_0$ → estimates differ significantly → use FE
- $p > 0.05$: cannot reject → RE is not detectably biased here →
  RE + ownership estimates are usable, with appropriate caveats

In most business/economics applications with firm- or country-level
data, the Hausman test rejects $H_0$ and FE is the right choice. The
RE + ownership estimates should then be treated as descriptive
(showing the between-unit pattern) rather than causal.

```{r}
#| label: final-re-table

modelsummary(
  list(
    "FE (main)"      = m_fe,
    "RE (main)"      = m_re,
    "RE + ownership" = m_re_own
  ),
  gof_map = c("nobs", "r.squared"),
  stars   = TRUE,
  notes   = c(
    "FE: within estimator, SEs clustered at firm level.",
    "RE: GLS (Swamy-Arora). HC robust SEs.",
    "Hausman test: report p-value here and preferred estimator."
  )
)
```

---

# First Differences

## The FD transformation

Both FE and FD eliminate $\eta_i$, but via different operations:

- **FE:** subtract each unit's time-mean $\bar{y}_i$
- **FD:** subtract the previous period's value $y_{i,t-1}$

Let us compute the first differences manually to see exactly what
the transformation does:

```{r}
#| label: fd-manual

wms_fd <- wms |>
  arrange(firmid, wave) |>
  group_by(firmid) |>
  mutate(
    d_management = management - lag(management),
    d_log_emp    = log_emp    - lag(log_emp),
    d_degree_m   = degree_m   - lag(degree_m)
  ) |>
  ungroup() |>
  filter(!is.na(d_management))   # first observation per firm has no lag

# How many differenced observations do we have?
nrow(wms_fd)

# Preview
wms_fd |>
  select(firmid, wave, management, d_management,
         log_emp, d_log_emp) |>
  head(10)
```

Running OLS on the differenced data gives the FD estimator directly:

```{r}
#| label: fd-ols

# OLS on demeaned data — this IS the FD estimator
m_fd_manual <- feols(d_management ~ d_log_emp + d_degree_m,
                     data = wms_fd)

m_fd_manual
```

`plm` handles this automatically:

```{r}
#| label: fd-plm

m_fd <- plm(management ~ log_emp + degree_m,
            data  = pdata,
            model = "fd")

m_fd
```

The two results should be identical (up to standard error convention).

## FE vs. FD side by side

```{r}
#| label: fe-fd-comparison

modelsummary(
  list(
    "Pooled OLS" = feols(management ~ log_emp + degree_m,
                         data = wms),
    "FE"         = m_fe,
    "FD"         = m_fd
  ),
  gof_map = c("nobs", "r.squared"),
  notes   = c(
    "FE: within estimator. FD: first-difference estimator.",
    "FE and FD are identical when T = 2 per unit."
  )
)
```

For WMS, most firms appear in only two waves — so FE and FD are
numerically very similar. The distinction matters most in longer panels.

## Interpreting FD differently from FE

The FD coefficient has a subtly different interpretation from FE:

**FE coefficient on `log_emp`:**
> For the same firm, when employment is above its own historical
> average, is management quality also above average?

**FD coefficient on `log_emp`:**
> Among firms that grew between consecutive survey waves,
> did management quality tend to improve over the same period?

Both are within-unit estimates, but FD is explicitly a
*change-on-change* comparison. This makes FD more natural when
the research question is framed in terms of change — for example,
evaluating a policy intervention or training programme.

## Visualising within-firm change

```{r}
#| label: fd-viz

# Track a handful of firms observed in multiple waves
multi_wave_firms <- wms |>
  count(firmid) |>
  filter(n >= 2) |>
  slice_sample(n = 8) |>
  pull(firmid)

wms |>
  filter(firmid %in% multi_wave_firms) |>
  ggplot(aes(x = wave, y = management,
             group = firmid, colour = factor(firmid))) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title   = "Management quality over time: eight firms",
    subtitle = "FD uses consecutive differences (arrows); FE uses deviations from the firm mean (dashed)",
    x       = "Survey wave",
    y       = "Management score (1–5)",
    colour  = "Firm ID"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

# Dynamic Panels: An Outlook

## When outcomes depend on their own past

Many management and economic processes are **persistent**: this
year's state is partly a function of last year's state.

In the WMS context: a firm that had strong management practices
in 2009 is likely to have strong practices in 2012 — because
practices do not change overnight, and because the people and
systems that produced the 2009 score are still in place.

The natural model adds $y_{i,t-1}$ as a regressor:

$$\text{management}_{it} = \alpha \cdot \text{management}_{i,t-1}
+ \beta \cdot x_{it} + \eta_i + v_{it}$$

$\alpha$ measures persistence: controlling for firm size, how
strongly does last wave's score predict this wave's?

## Why FE produces biased estimates here

Applying the within-transformation to a dynamic model creates
a mechanical problem — the **Nickell bias**.

The demeaned lagged outcome $(y_{i,t-1} - \bar{y}_i)$ is
*correlated with* the demeaned error $(v_{it} - \bar{v}_i)$
because $\bar{v}_i$ contains $v_{i,t-1}$, which is correlated
with $y_{i,t-1}$.

**The result:** FE systematically underestimates $\alpha$. The bias
does not vanish unless $T \to \infty$, and for panels with small
$T$ — like the WMS — it can be severe.

```{r}
#| label: nickell-illustration

# Pooled OLS: biased upward (includes between-firm heterogeneity)
m_dyn_ols <- feols(management ~ lag(management) + log_emp,
                   data = wms)

# FE: Nickell bias — biased downward
m_dyn_fe  <- feols(management ~ lag(management) + log_emp | firmid,
                   vcov = ~firmid,
                   data = wms)

modelsummary(
  list("Dynamic OLS" = m_dyn_ols, "Dynamic FE (biased)" = m_dyn_fe),
  gof_map = c("nobs", "r.squared", "r2.within"),
  notes   = c(
    "OLS biased upward; FE biased downward (Nickell).",
    "True persistence parameter lies between these estimates.",
    "Do not interpret the FE coefficient causally."
  )
)
```

The OLS estimate is pushed up by unobserved firm quality; the FE
estimate is pulled down by the Nickell bias. The true persistence
lies somewhere between — but neither estimate is reliable.

## The solution: Arellano-Bond GMM

The standard solution is the **Arellano-Bond estimator** (1991),
also called Diff-GMM:

1. **First-difference** to eliminate $\eta_i$
2. **Instrument** $\Delta y_{i,t-1}$ with lagged *levels* of $y$
   ($y_{i,t-2}, y_{i,t-3}, \ldots$) — they are correlated with
   the differenced lag but, under no serial correlation in $v_{it}$,
   uncorrelated with the differenced error

An extension, **System-GMM** (Blundell & Bond, 1998), uses both the
differenced and levels equations for better efficiency when the
outcome is highly persistent.

::: {.callout-note}
## For further study
Dynamic panel GMM is technically rich and takes time to do properly.
The essential practical reference is:

Roodman, D. (2009). How to do xtabond2: An introduction to Diff and
Sys GMM in Stata. *Stata Journal*, 9(1), 86–136.
The logic translates directly to `plm::pgmm()` in R.

Recognising the *problem* (Nickell bias) is the critical skill for
now — knowing when you need GMM is more important than knowing how
to run it without understanding it.
:::

---

# Quarto Skill: Documenting Estimator Choice

When reporting results from multiple estimators, readers need to
understand *why* you chose the one you did. Three things to always
include:

**1. State the estimator and SE approach in table notes:**

```{r}
#| label: quarto-skill

modelsummary(
  list("FE" = m_fe, "RE" = m_re, "FD" = m_fd),
  gof_map = c("nobs", "r.squared"),
  stars   = TRUE,
  notes   = c(
    "FE: within estimator, SEs clustered at firm level.",
    "RE: GLS (Swamy-Arora). FD: first-difference, HC robust SEs.",
    "Hausman test rejects RE (p < 0.05); FE preferred for causal inference.",
    "FD shown for robustness; equivalent to FE when T = 2."
  )
)
```

**2. Reference the specification test that guided your choice.**
A single sentence — "the Hausman test rejects RE ($p = 0.02$);
we therefore use FE as the main specification" — is sufficient.

**3. Show the robustness column.** Including FD or RE alongside
your main FE result shows that conclusions are not sensitive
to the choice of estimator. Stability across columns builds
credibility.

---

# Summary

**Random Effects:** treats $\eta_i$ as random; consistent only when
$\eta_i \perp x_{it}$; can estimate time-invariant covariates;
always test with `plm::phtest()` before preferring RE over FE.

**First Differences:** subtracts $y_{i,t-1}$ from $y_{it}$;
eliminates $\eta_i$; natural for change questions; equivalent
to FE when $T = 2$; more robust when errors follow a random walk.

**Dynamic panels:** including $y_{i,t-1}$ as a regressor makes FE
inconsistent (Nickell bias); Arellano-Bond GMM is the solution;
recognise the problem even if you do not yet implement the fix.

**In R:** `plm::plm(..., model = "within" / "random" / "fd")`;
`plm::phtest()` for the Hausman test.

---

# Further Reading

- Baltagi, B.H. (2021). *Econometric Analysis of Panel Data* (6th ed.).
  Springer. The standard reference for all panel methods covered here.
- Croissant, Y. & Millo, G. (2008). Panel data econometrics in R:
  The plm package. *Journal of Statistical Software*, 27(2).
  Essential reading for R implementation — covers RE, FD, and tests.
- Roodman, D. (2009). How to do xtabond2. *Stata Journal*, 9(1).
  The practical guide to dynamic panel GMM.
- Huntington-Klein, N. (2022). *The Effect*. Ch. 16–17.
  Available at [theeffectbook.net](https://theeffectbook.net)
