---
title: "Exercise: Advanced Panel Data"
subtitle: "Session 6b · Quantitative Data Analysis: Applied Data Science and Modern Econometrics"
author: "Your name here"
date: "Optional — see course website for scheduling"
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note}
## Business question
*Does ownership type explain differences in management quality —
and do firms that grow also improve their management practices?*

This exercise builds on Session 6. You now have three estimators
in your toolkit — FE, RE, and FD — and the Hausman test to choose
between the first two. Work through the tasks in order. Write your R
code in the chunks provided and your interpretation in the text below.

You have approximately **30 minutes**.
:::

---

# Setup

Run this chunk first.

```{r}
#| label: setup
library(tidyverse)
library(fixest)
library(plm)
library(modelsummary)

wms <- read_csv("data/wms_da_textbook.csv") |>
  mutate(log_emp = log(emp_firm))

pdata <- pdata.frame(wms, index = c("firmid", "wave"))
```

---

# Task 1 — Is ownership type time-invariant?

Before choosing between FE and RE, check whether `ownership` actually
varies within firms across waves. If it does not change, FE cannot
estimate its effect.

```{r}
#| label: task1-check
# Count how many distinct ownership types each firm has across waves
# Hint: group_by(firmid) |> summarise(n_types = n_distinct(ownership))
# Then count how many firms have n_types == 1 vs > 1
```

**Write here:** Is `ownership` effectively time-invariant in this
dataset? What does this imply for whether FE can estimate its effect?

---

# Task 2 — RE and the ownership effect

Estimate three models using `plm()` on `pdata`:

- `m_fe`: FE (`model = "within"`) with `log_emp` and `degree_m`
- `m_re`: RE (`model = "random"`) with `log_emp` and `degree_m`
- `m_re_own`: RE with `log_emp`, `degree_m`, and `ownership`

```{r}
#| label: task2-models
# m_fe     <- plm(management ~ log_emp + degree_m, data = pdata, model = "within")
# m_re     <- plm(management ~ log_emp + degree_m, data = pdata, model = "random")
# m_re_own <- plm(management ~ log_emp + degree_m + ownership, ...)
```

Present all three in a single `modelsummary()` table.

```{r}
#| label: task2-table
# Your table here
```

**Interpret:**

1. Do the `log_emp` and `degree_m` coefficients differ much between
   FE and RE? What does this tell you about the role of between-unit
   variation in these estimates?

2. Looking at the RE + ownership model: which ownership types predict
   higher management quality? Does this match your prior expectations?

---

# Task 3 — The Hausman test

Run the Hausman test to assess whether the RE model is defensible.

```{r}
#| label: task3-hausman
# phtest(m_fe, m_re)
```

**Interpret:**

1. State the null hypothesis in plain English — what is the test
   actually checking?

2. What does your test result say? Should you prefer FE or RE as
   your main specification?

3. Given this result, how should you treat the ownership coefficients
   from Task 2? Are they a reliable causal estimate or a descriptive
   pattern?

---

# Task 4 — First Differences

Estimate a **First Differences** model of `management` on
`log_emp` and `degree_m` using `plm(..., model = "fd")`.

```{r}
#| label: task4-fd
# m_fd <- plm(management ~ log_emp + degree_m, data = pdata, model = "fd")
```

Present pooled OLS, FE, and FD in a single table.

```{r}
#| label: task4-table
# Your table here
# For pooled OLS: feols(management ~ log_emp + degree_m, data = wms)
```

**Interpret:**

1. Rephrase the FD coefficient on `log_emp` in one sentence, using
   the change-on-change language appropriate for FD. Begin with:
   *"Among firms that grew between survey waves..."*

2. FE and FD give similar estimates here. Why would you expect that
   given the structure of the WMS panel?

---

# Task 5 — Reflection and decision

::: {.callout-warning}
## Think before you code
This task requires no new code. Write 4–6 sentences in total.
:::

You have now estimated the management-size relationship using
four estimators: pooled OLS, FE, RE, and FD.

**Answer the following:**

1. A colleague argues: "RE is more efficient than FE, so we should
   always prefer it." Write two sentences explaining why this
   argument is incomplete and under what conditions it would be
   wrong in this specific dataset.

2. You are writing a thesis section that asks: "Did firms that
   expanded their workforce between 2009 and 2015 also improve
   their management practices?" Which estimator from today is
   most naturally suited to this question, and why?

---

# Optional Extension

::: {.callout-tip}
## If you finish early
The demo showed that including `lag(management)` in a FE model
produces biased estimates (Nickell bias). Try it yourself:

Estimate two dynamic models using `feols()`: one pooled OLS and
one FE, both including `lag(management)` and `log_emp`. Compare
the coefficient on `lag(management)` across the two. Based on the
theory of Nickell bias, in which direction would you expect the
FE estimate to be biased relative to the true value? Does the
comparison with OLS support this?
:::

```{r}
#| label: optional-dynamic
# m_dyn_ols <- feols(management ~ lag(management) + log_emp, data = wms)
# m_dyn_fe  <- feols(management ~ lag(management) + log_emp | firmid,
#                    vcov = ~firmid, data = wms)
```
