---
title: "Advanced Panel Data"
subtitle: "Session 6b (Optional) — Quantitative Data Analysis: Applied Data Science and Modern Econometrics"
author: "Claudius Gräbner-Radkowitsch"
institute: "Europa-Universität Flensburg & Johannes Kepler University Linz"
date: "Optional — see course website for scheduling"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    progress: true
    code-line-numbers: false
    highlight-style: github
    fig-align: center
    incremental: false
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
#| eval: true
library(tidyverse)
library(fixest)
library(plm)
library(modelsummary)
```

# Where We Are {background-color="#2C5F8A"}

## Session 6 recap

In Session 6 we learned that panel data contain two sources of
variation — between units and within units over time.

The **Fixed Effects (FE) estimator** exploits only within-unit
variation by subtracting each unit's time-mean. This eliminates
stable unobserved heterogeneity $\eta_i$ and corrects the OVB
that pooled OLS suffers from.

**Two-way FE** additionally controls for common period shocks $\gamma_t$.

::: {.callout-note}
## Today's question
FE is powerful — but it rests on assumptions that sometimes fail.
What do we do when they do?
:::

## Three situations FE does not handle well

```{=html}
<div style="display: grid; grid-template-columns: 1fr 1fr 1fr; gap: 1.5rem; margin-top: 1rem; font-size: 0.85em;">
  <div style="background: #f0f4f8; border-radius: 8px; padding: 1.2rem;">
    <p style="font-weight: bold; color: #2C5F8A; margin-bottom: 0.5rem;">1. Time-invariant covariates matter</p>
    <p>FE differences them out — they cannot be estimated.</p>
    <p style="color: #555; margin-top: 0.5rem;">→ <strong>Random Effects</strong></p>
  </div>
  <div style="background: #f0f4f8; border-radius: 8px; padding: 1.2rem;">
    <p style="font-weight: bold; color: #2C5F8A; margin-bottom: 0.5rem;">2. The question is about change</p>
    <p>FE uses deviations from means — not consecutive differences.</p>
    <p style="color: #555; margin-top: 0.5rem;">→ <strong>First Differences</strong></p>
  </div>
  <div style="background: #f0f4f8; border-radius: 8px; padding: 1.2rem;">
    <p style="font-weight: bold; color: #2C5F8A; margin-bottom: 0.5rem;">3. The outcome is dynamic</p>
    <p>Lagged $y_{i,t-1}$ as regressor — FE is biased here.</p>
    <p style="color: #555; margin-top: 0.5rem;">→ <strong>Dynamic panel models</strong></p>
  </div>
</div>
```

<br>

Same dataset throughout: `wms-management-survey`.

---

# Random Effects {background-color="#2C5F8A"}

## FE throws away between-unit information

The within-transformation removes everything that does not vary
over time within a unit — including covariates we care about.

**In the WMS dataset:**

- `ownership` type (founder, family, dispersed) is largely stable
  within firms across survey waves
- FE silently drops it — its coefficient cannot be estimated

To recover estimates for time-invariant covariates, we need a
model that retains between-unit information.

## The RE model

**Random Effects (RE)** treats $\eta_i$ not as a fixed unknown
parameter per unit, but as a random draw from a distribution:

$$\eta_i \sim \mathcal{N}(0, \sigma^2_\eta)$$

Because $\eta_i$ is modelled as part of the error structure rather
than differenced out, time-invariant covariates remain in the model
and can be estimated.

Estimation proceeds via **Generalised Least Squares (GLS)** —
a weighted average of the within and between estimators.

::: {.callout-warning}
## The critical assumption
RE is only consistent when $\eta_i$ is **uncorrelated** with all
regressors. If firms with strong management culture (high $\eta_i$)
also tend to be larger — which is plausible — RE is biased.
:::

## The efficiency tradeoff

```{=html}
<div style="display: grid; grid-template-columns: 1fr 1fr; gap: 2rem; margin-top: 0.5rem; font-size: 0.88em;">
  <div style="border: 2px solid #2C5F8A; border-radius: 8px; padding: 1.2rem;">
    <p style="font-weight: bold; color: #2C5F8A; margin-bottom: 0.5rem;">Fixed Effects</p>
    <p>✓ Consistent even if $\eta_i$ correlates with $x_{it}$</p>
    <p>✓ Safe default in business/economics</p>
    <p>✗ Cannot estimate time-invariant covariates</p>
    <p>✗ Less efficient when RE assumption holds</p>
  </div>
  <div style="border: 2px solid #4A9B6F; border-radius: 8px; padding: 1.2rem;">
    <p style="font-weight: bold; color: #4A9B6F; margin-bottom: 0.5rem;">Random Effects</p>
    <p>✓ Can estimate time-invariant covariates</p>
    <p>✓ More efficient when $\eta_i \perp x_{it}$</p>
    <p>✗ Biased if $\eta_i$ correlates with $x_{it}$</p>
    <p>✗ Stronger distributional assumption required</p>
  </div>
</div>
```

<br>

**When RE is preferred:** units are a random sample from a larger
population *and* time-invariant covariates are of substantive
interest — more common in individual surveys than in firm panels.

## The Hausman test: choosing between FE and RE

**The question:** is $\eta_i$ correlated with $x_{it}$?

- If **no**: FE and RE both consistent; RE is more efficient
- If **yes**: RE is biased; FE is the right choice

The **Hausman test** compares the two sets of estimates. Under
$H_0$ (no correlation), they should be similar. A significant
divergence is evidence that RE is inconsistent.

```{r}
library(plm)

pdata <- pdata.frame(wms, index = c("firmid", "wave"))

m_fe <- plm(management ~ log_emp + degree_m,
            data = pdata, model = "within")
m_re <- plm(management ~ log_emp + degree_m,
            data = pdata, model = "random")

phtest(m_fe, m_re)
```

::: {.callout-tip}
## Rule of thumb
Reject $H_0$ → use FE. In most business/economics panel
applications, the Hausman test rejects and FE is appropriate.
:::

## RE in practice: estimating ownership effects

```{r}
# RE without ownership — for Hausman comparison
m_re <- plm(management ~ log_emp + degree_m,
            data = pdata, model = "random")

# RE with ownership — the quantity FE cannot give us
m_re_own <- plm(management ~ log_emp + degree_m + ownership,
                data = pdata, model = "random")

modelsummary(
  list("FE" = m_fe, "RE" = m_re, "RE + ownership" = m_re_own),
  gof_map = c("nobs", "r.squared"),
  notes   = "FE: within estimator. RE: GLS (Swamy-Arora)."
)
```

::: {.callout-note}
## Interpretation
If the Hausman test rejects, treat RE + ownership estimates with
caution — the ownership coefficient absorbs some of the
between-unit bias. Report the Hausman result alongside.
:::

---

# First Differences {background-color="#2C5F8A"}

## An alternative to the within-transformation

Both FE and FD eliminate $\eta_i$ — but via different operations.

**FE:** subtract each unit's *time-mean*
$$\tilde{y}_{it} = y_{it} - \bar{y}_i$$

**FD:** subtract the *previous period's value*
$$\Delta y_{it} = y_{it} - y_{i,t-1}$$

Both remove $\eta_i$ because it is time-invariant. One observation
per unit is lost with FD (no lag for $t = 1$).

## FD vs. FE: when does the choice matter?

**When $T = 2$:** FD and FE are numerically identical.

**When $T > 2$:** they differ — and the right choice depends on
the error structure.

- **FE is more efficient** when $v_{it}$ is serially uncorrelated
- **FD is more robust** when $v_{it}$ follows a random walk —
  FE amplifies that correlation, FD removes it

::: {.callout-note}
## The natural use case for FD
FD is most intuitive when the research question is explicitly
about **change**: did firms that *improved* their management
score also *improve* performance? The differenced data directly
represent the changes of interest.
:::

## FD in R

```{r}
m_fd <- plm(management ~ log_emp + degree_m,
            data  = pdata,
            model = "fd")

modelsummary(
  list("Pooled OLS" = feols(management ~ log_emp + degree_m,
                            data = wms),
       "FE"         = m_fe,
       "FD"         = m_fd),
  gof_map = c("nobs", "r.squared"),
  notes   = "FE: within estimator. FD: first-difference estimator."
)
```

**Rephrasing the FD coefficient:**

> Among firms that grew between survey waves, did management quality
> tend to improve over the same period?

This is a change-on-change interpretation — distinct from FE's
deviation-from-mean interpretation.

---

# Dynamic Panels: An Outlook {background-color="#2C5F8A"}

## When outcomes persist over time

Many business processes are **path-dependent**: today's outcome
is partly determined by yesterday's.

- Management quality: practices change slowly — last wave's score
  is a strong predictor of this wave's
- Firm performance: last year's profit shapes this year's investment
- Employee capabilities: skills accumulate over time

The natural model adds $y_{i,t-1}$ as a regressor:

$$y_{it} = \alpha \, y_{i,t-1} + x_{it}\beta + \eta_i + v_{it}$$

$\alpha$ measures **persistence**: how strongly does last period's
outcome predict this period's, holding $x_{it}$ constant?

## Why FE fails here: the Nickell bias

Applying the within-transformation to a dynamic model creates a
mechanical problem.

The demeaned lagged outcome $(y_{i,t-1} - \bar{y}_i)$ is
**correlated with** the demeaned error $(v_{it} - \bar{v}_i)$
because $\bar{v}_i$ contains $v_{i,t-1}$, which is correlated
with $y_{i,t-1}$.

**Consequence:** the FE estimate of $\alpha$ is biased downward.
The bias disappears only as $T \to \infty$ — for panels with
small $T$ it can be severe.

::: {.callout-warning}
## A practical warning
If your model includes a lagged dependent variable and your panel
has small $T$: FE is inconsistent. The model *looks* like it runs,
but the estimates are unreliable.
:::

## The solution: dynamic panel GMM

The standard solution is the **Arellano-Bond estimator**
(Diff-GMM), which:

1. First-differences the equation to remove $\eta_i$
2. Uses lagged *levels* of $y$ as instruments for the differenced
   lagged term — they are correlated with $\Delta y_{i,t-1}$ but
   not with $\Delta v_{it}$ (under no serial correlation in $v_{it}$)

An extension, **System-GMM** (Blundell-Bond), also uses the levels
equation alongside the differenced equation for better efficiency
when outcomes are highly persistent.

::: {.callout-note}
## For further study
GMM estimation for dynamic panels is a rich and technically
demanding topic. The practical reference is Roodman (2009),
*How to do xtabond2*, available at
[ideas.repec.org](https://ideas.repec.org). In R, the `pgmm()`
function in the `plm` package implements both Diff-GMM and Sys-GMM.
:::

---

# The Decision Landscape {background-color="#2C5F8A"}

## Choosing an estimator

| Situation | Recommended estimator |
|---|---|
| Static; $\eta_i$ likely correlated with $x$ | **Unit FE** ← default |
| Static; time-invariant covariates matter; Hausman not rejected | **Random Effects** |
| Question is explicitly about *change* | **First Differences** |
| Dynamic ($y_{i,t-1}$ as regressor); small $T$ | **Diff-GMM / Sys-GMM** |
| Large $T$; potential non-stationarity | Long panel / cointegration |

When uncertain: **two-way FE is the safe starting point**.

## Common pitfalls

- **Pooled OLS without checking:** always run a Hausman test or
  compare FE results before reporting pooled estimates

- **Ignoring clustering:** panel standard errors without unit-level
  clustering are almost always too small

- **FE with a lagged DV:** if $y_{i,t-1}$ is in the model and
  $T$ is small, FE is biased — recognise the dynamic panel setting

- **RE without the Hausman test:** never prefer RE on efficiency
  grounds alone without testing the exogeneity assumption

- **Confusing within and overall $R^2$:** always report within-$R^2$
  for FE models

---

# Session Summary {background-color="#2C5F8A"}

## What we covered today

**Random Effects:** treats $\eta_i$ as random; consistent only when
$\eta_i \perp x_{it}$; can estimate time-invariant covariates;
choose using the Hausman test (`plm::phtest()`).

**First Differences:** subtracts $y_{i,t-1}$ from $y_{it}$;
eliminates $\eta_i$; natural for change questions; equivalent to
FE when $T = 2$; more robust when errors follow a random walk.

**Dynamic panels (outlook):** Nickell bias makes FE inconsistent
when $y_{i,t-1}$ is a regressor; Arellano-Bond GMM is the standard
solution; beyond this course but important to recognise.

**Decision map:** two-way FE is the default; move to RE when
time-invariant covariates matter and Hausman allows it; use FD
when the question is about change; seek GMM only when dynamics
are substantively important.

## Quarto skill: documenting estimator choice

When reporting results from multiple estimators, always explain
*why* you chose the one you did:

```{r}
modelsummary(
  list("FE" = m_fe, "RE" = m_re, "FD" = m_fd),
  gof_map = c("nobs", "r.squared"),
  notes   = c(
    "FE: within estimator, SEs clustered at firm level.",
    "RE: GLS (Swamy-Arora). FD: first-difference, HC robust SEs.",
    "Hausman test rejects RE (p < 0.05); FE preferred for causal inference.",
    "FD included for robustness — identical to FE when T = 2."
  )
)
```

::: {.callout-tip}
## Good practice
One footnote sentence stating the Hausman test result and your
estimator choice is standard in applied panel econometrics.
:::
